---
title: "Common Functions"
author: "João Neto"
date: October 2014
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 6
cache: yes
---

NA removal
----------
```{r}
l5 <- c(1,2,NA,4,NA,6)
l5a <- l5[!is.na(l5)]
l5a
l5b <- c("a","b",NA,"d",NA,"f") 
good.pairs <- complete.cases(l5,l5b) # if we need to join both lists only with complete pairs...
data.frame(a=l5[good.pairs],b=l5b[good.pairs]) # another example: airquality is a data frame
airquality[4,][c(1,4,3)] # shows cols 1, 4 & 3 from the 4th row
airquality[complete.cases(airquality),][1:6,] # to show all non NA rows for the first 6 columns
```

Sample function
---------------

Function **sample** takes a sample of the specified size from the elements of x using either with or without replacement.

```{r}
coin <- c("Heads","Tails")
sample(coin,8,rep=T)
dice.events <- 1:6  # dice possible results
ten.throws <- sample(dice.events, 10, replace=T)  # balanced dice
ten.throws
ten.more.throws <- sample(dice.events, 10, replace=T, prob=c(.1,.1,.1,.1,.1,.5))  # unbalanced dice (50% it falls 6)
ten.more.throws
# Without replacement, we can create permutations
xs <- 1:10
sample(xs,10)
sample(xs,10)
# Eg: we have 50 pacients that we want to separate into two equal size groups (Control or Treatment)
pacients <- data.frame(patient = 1:50,
                       age = rnorm(100, mean = 60, sd = 12))
head(pacients)
xs <- rep(c("Control", "Treat"),25)
xs
xs <- sample(xs,50)  # a random permutation to remove selection bias
xs
pacients$group <- as.factor(xs)  # add new column with required information
head(pacients)
# Eg: simple bootstrap example (from: www.sigmafield.org/2010/05/23/r-function-of-the-day-sample)
rn <- rnorm(1000, 10) # random sample from a normal distribution
# making 1000 samples and keeping the means
sample.means <- replicate(1000, mean(sample(rn, replace = TRUE)))
sample.means[1:20]
quantile(sample.means, probs = c(0.025, 0.975)) # produces sample quantiles corresponding to the given probabilities
# Compare this to the standard parametric technique.
t.test(rn)$conf.int
```

split-apply-combine
---------------

References:
+ [http://www.sigmafield.org/2009/09/20/r-function-of-the-day-tapply](http://www.sigmafield.org/2009/09/20/r-function-of-the-day-tapply)
+ [http://stackoverflow.com/questions/11562656/averaging-column-values-for-specific-sections-of-data-corresponding-to-other-col/11562850#11562850](http://stackoverflow.com/questions/11562656/averaging-column-values-for-specific-sections-of-data-corresponding-to-other-col/11562850#11562850)

check also: [http://www.jstatsoft.org/v40/i01/paper](http://www.jstatsoft.org/v40/i01/paper)

We use **tapply** when:
* A dataset that can be broken up into groups
* We want to break it up into groups
* Within each group, we want to apply a function

The tapply function is useful when we need to break up a vector into groups defined by some classifying factor, compute a function on the subsets, and return the results in a convenient form.

```{r}
## generate data for medical example
medical.example <- data.frame(patient = 1:100,
        age = rnorm(100, mean = 60, sd = 12),
        treatment = gl(2, 50, labels = c("Treatment", "Control")))
# we want to break the dataset by Treatment, and then find the age's mean 
tapply(medical.example$age, medical.example$treatment, mean)
# Another eg
baseball.example <- data.frame(team = gl(5, 5,
    labels = paste("Team", LETTERS[1:5])),
    player = sample(letters, 25), batting.average = runif(25, .200, .400))
# we want to break the dataset by Team, and then find the max batting avg
tapply(baseball.example$batting.average, baseball.example$team, max)
# an artificial eg from R's help:
n <- 17
fac <- factor(rep(1:3, length = n), levels = 1:5)
fac
table(fac)
# this next function, separates all 1:n value into the factors given by fac, 
# and then sum each factor
tapply(1:n, fac, sum)
# Let's explicitly sum the first factor just to confirm:
fac == 1
(1:n)[fac==1]
sum((1:n)[fac==1])
# we can apply any unary function over each factor
tapply(1:n, fac, range)
tapply(1:n, fac, quantile)
```

Other possibilities:

```{r}
df <- data.frame(dive=factor(sample(c("dive1","dive2"),10,replace=TRUE)),
                 speed=runif(10))
df

# 'by()' takes in vectors and applies a function to them
res.by <- by( df$speed, df$dive, mean)
library(taRifx)
as.data.frame(res.by)

# takes in data.frames, outputs data.frames, and uses a formula interface.
aggregate( speed ~ dive, df, mean )

# Check also library plyr which facilitates this type of data frama manipulation
```


Rle function
---------------

Function **rle** computes the lengths and values of runs of equal values in a vector. 

Function **inverse.rle** does  the reverse operation.

```{r}
coin <- c("H","T")
experiment <- sample(coin,100,rep=T)
experiment
experiment.rle <- rle(experiment)
max.seq <- max(experiment.rle$lengths) # the maximum sequence of equal values
max.seq
experiment.rle$values[which(experiment.rle$lengths == max.seq)] # was it heads or tails?
# we can compute both max sequences of heads and tails by using tapply:
tapply(experiment.rle$lengths, experiment.rle$values, max)
inverse.rle(experiment.rle) #  recomputes the initial sequence
```


Split/Cut functions
--------------

Function **split** divides the data in the vector x into the groups defined by f. The replacement forms replace values corresponding to such a division.

Function **cut** divides the range of x into intervals and codes the values in x according to which interval they fall. The leftmost interval corresponds to level one, the next leftmost to level two and so on.

```{r}
split(1:10, 1:2)
split(1:10, 1:2)[[1]]  # or: split(1:10, 1:2)$"1"
ma <- cbind(x = 1:10, y = (-4:5)^2)
ma
split(ma, colnames(ma))
split(ma, col(ma))
split(ma, col(ma))[[1]] 
ma[,1]                   # easier alternative
ma[,"x"]                 # another alternative
xs <- 1:20
split(xs,factor(c("yes","no","yes","yes"))) # every four elements are attached to the given level sequence "yes","no","yes","yes"
is <- split(xs,factor(c("yes","no","yes","yes")))[["yes"]]
xs[is]
# Use of cut (from: www.sigmafield.org/2009/09/23/r-function-of-the-day-cut)
# generate some data 
clinical.trial <-
    data.frame(patient = 1:100, age = rnorm(100, mean = 60, sd = 8),
               year = sample(paste("19", 85:99, sep = ""), 100, replace = T))
# Now, we will use the cut function to make age a factor, ie, 
# age will be a categorical variable
c1 <- cut(clinical.trial$age, breaks = 4)
table(c1)
# year.enroll is a factor, so must convert to numeric first!
years <- as.numeric(as.character(clinical.trial$year))
years
c2 <- cut(years, breaks = 3)
table(c2)
# The previous intervals where determined by R. 
# We can force our own (in this case, I'll use 'seq')
c3 <- cut(clinical.trial$age, breaks = seq(30, 80, by = 10))
table(c3)
# the interval size can also be variable:
age.cat <- function(x, lower = 0, upper, by = 10,
                   sep = "-", above.char = "+") {

 labs <- c(paste(seq(lower, upper - by, by = by),
                 seq(lower + by - 1, upper - 1, by = by),
                 sep = sep),
           paste(upper, above.char, sep = ""))

 cut(floor(x), breaks = c(seq(lower, upper, by = by), Inf),
     right = FALSE, labels = labs)
}
# This function categorizes age in a fairly flexible way. The first assignment to labs inside the function creates a vector of labels. Then, the cut function is called to do the work, with the custom labels as an argument.

# only specifying an upper bound, uses 0 as lower bound, and breaks up categories by 10
table(age.cat(clinical.trial$age, upper = 70))
# now specifying a lower bound
table(age.cat(clinical.trial$age, lower = 30, upper = 70))
# now specifying a lower bound AND the "by" argument 
table(age.cat(clinical.trial$age, lower = 30, upper = 70, by = 5))
```

Subsetting
-----------

`subset` returns subsets of vectors, matrices or data frames which meet conditions.

```{r}
head(airquality)

head(subset(airquality, Temp > 80, select = c(Ozone, Temp)))
head(subset(airquality, Day == 1, select = -Temp))
head(subset(airquality, select = Ozone:Wind))
```

