---
title: "Expressions"
author: "João Neto"
date: October 2014
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 6
cache: yes
---

Introduction
--------------

In R a **call** is an unevaluated expression which consists of the named function applied to the given arguments. Function `call` is used to create a call. It receives a function's name, plus a series of arguments to be applied to the function. `eval` is used to evaluate the final result.

```{r}
a <- 25
b <- call("sqrt", a)
b
eval(b)
a <- 16               # does not influence the previous environment
eval(b)
is.call(b)
is.call(call)         # functions are not calls
c <- call("^", 2, 4)  # call can receive multiple arguments
eval(c)
```

`quote` returns the argument as non evaluated:

```{r}
a <- 25
b <- call("sqrt", quote(a))
eval(b)
a <- 16               # now it influences, since R still not evaluated the parameter
eval(b)  
eval(quote(b), env=list(b=1)) # reads from an environment (can also be a list or a dataframe)
```

`do.call` constructs and executes a function call from a name or a function and a list of arguments to be passed to it.

```{r}
a <- 10
b <- 2
f <- function (a,b) a/b
do.call("f", args=list(a+1, b))
do.call("f", args=list(b=a, a=b))
# make an environment
env <- new.env()
assign("a", 2, envir = env) # same as env$a <- 2
assign("b", 8, envir = env)
assign("f", function(a,b) a+b, envir = env)
as.list(env)
do.call("f", args=list(quote(a), quote(b)), envir=env)
env$f <- function(a,b) a^b
do.call("f", args=list(quote(a), quote(b)), envir=env)
```

`substitute` return the unevaluated expression, replacing any variables bound in the environment. The environment is a given list of assignments, or if omitted is the current evaluation environment.


```{r}
substitute(a+b, list(a=1))
a <- substitute(a+b+c, list(a=1,c=5))
a
eval(a, list(b=10))
```

We can use `eval` and `substitute` to implement `subset`, a function that return subsets of vectors, matrices or data frames which meet conditions:

```{r}
df <- data.frame(x=11:18, y=18:11)
df
subset(df, x>y)

my.subset <- function(x, condition) {
  condition_call <- substitute(condition)
  rows <- eval(condition_call, env=x, enclos=parent.frame()) # parent.frame is the var scope the user needs
  x[rows, ]
}

my.subset(df, x>y)
a <- 15
my.subset(df, x>a)
  
```

Notice that these type of functions are no longer referentially transparent. A function is referentially transparent if you can replace its arguments with their values and its behaviour doesn't change. For example, if a function, f(), is referentially transparent and both x and y are 10, then f(x), f(y), and f(10) will all return the same result. Check [here](http://adv-r.had.co.nz/Computing-on-the-language.html) for more info

Expressions
----------

Expressions are calls in R. Function `expression` returns a vector of type "expression" containing its arguments (unevaluated).

```{r}
expr <- expression(x^2 + b*x)
is.expression(expr)
eval(expr, list(x=10, b=3))
eval(expr, list(x=c(10,20,30), b=1:3))
```

`all.vars` return a character vector containing all the names which occur in an expression or call.

```{r}
expr <- expression(x^2 + b*x)
all.vars(expr)
all.vars(quote(expr))
```


Symbolic Computation
----------------

We can do some basic symbolic computation. `D` and `deriv` compute the derivate of an expression:

```{r}
expr <- expression(x^2 + b*x)
de.dx <- D(expr, "x")   # for simple variable
de.dx
de.db <- D(expr, "b")
de.db
eval(de.dx, list(b=1, x=10))

# computing n-th derivative
# pre: n>0
Dn <- function(expr, name, n=1) {
   if (n == 1)
     D(expr, name)
   else
     Dn(D(expr, name), name, n-1)
}
Dn(expression(sin(x^2)), "x", 3)

expr <- expression(x^2 + b*x*y + y^3)
deriv(expr, namevec=c("x","y"))               # for multiple variables
d <- deriv(expr, namevec=c("x","y"), hessian=TRUE) # includes the Hessian, ie, the matrix of second derivatives 
d
eval(d, list(x=1,y=3))
```

A more complex eg
------------------

[Ref](http://oddhypothesis.blogspot.pt/2014/08/optimizing-with-r-expressions.html). We are trying to fit data $x$ into the nonlinear model:

$$\frac{K y_0 e^{u(x-tl)}}{K + y_0(e^{u(x-tl)}-1)} + b_1 + (b_0-b_1)e^{-kx} + b_2x$$

```{r}
# the model equation
expr <- expression( (K*y0*exp(u*(x-tl)))/(K + y0*(exp(u*(x-tl))-1)) + 
                    b1 + (b0 - b1)*exp(-k*x) + b2*x )

all.vars(expr)
```

The next line produces a list of the partial derivatives of the above equation with respect to each parameter:

```{r}
ds <- sapply(all.vars(expr), function(v) {D(expr, v)} )
ds
class(ds)
class(ds[[1]])
```

Each element of this list is itself an expression.

Now, if we assign values to the parameters, we can compute the Jacobian matrix $J$, necessary to compute the gradient:

```{r}
jacob <- function(expr, env) {
   t( sapply(all.vars(expr), function(v) {eval(D(expr, v), env=env)} ) )
}
```

So, let's give them some values:

```{r}
# this will be the environment for the evaluation of J
ps <- c(y0=0.01, u=0.3, tl=5, K=2, b0=0.01, b1=1, b2=0.001, k=0.1)
x  <- seq(0,10)

J <- jacob(expr, env= c(as.list(ps), list(x=x)))
J <- J[names(ps),,drop=F]  # drop 'x' row which refers to the independent variable
J
```

The Hessian $H$ is approximately $H \approx J^TJ$

```{r}
H <- J %*% t(J)  # because linear algebra in R is a little strange, the transpose is applied to the 2nd Jacobian
```

The gradient is $g = -J r$ where $r$ are the residuals.

We can box all this into a class:

```{r}
ModelObject = setRefClass('ModelObject', 
                          
  fields = list(
    name = 'character',
    expr = 'expression'
  ),
  
  methods = list(
    value = function(p, data){
      eval(.self$expr, c(as.list(p), as.list(data)))
    },
  
    jacobian = function(p, data){
      J = t(sapply(all.vars(.self$expr), function(v, p, data){
              eval(D(.self$expr, v), c(as.list(p), as.list(data)))
            }, p=p, data=data))

      return(J[names(p),,drop=F])
    },
    
    gradient = function(p, data){
        r = data$y - value(p, data)
        return(-jacobian(p, data) %*% r)
    },
    
    hessian = function(p, data){
      J = jacobian(p, data)
      return(J %*% t(J))
    }
  )
)
```

So let's make some fake data and test the model:

```{r}
# the model expression
expr <- expression( (K*y0*exp(u*(x-tl)))/(K + y0*(exp(u*(x-tl))-1)) + 
                    b1 + (b0 - b1)*exp(-k*x) + b2*x )

# make some data:
xs <- seq(0,48,by=0.25)
p0 <- c(y0=0.01, u=0.3, tl=5, K=2, b0=0.01, b1=1, b2=0.001, k=0.1)  # true values of the parameters
xy <- list(x=xs,
           y=eval(expr, envir=c(as.list(p0), list(x=xs)))
          )
  
plot(xy, main='Fit Results', type="l", col="blue", lty=2, lwd=2); # target function
xy$y <- xy$y+rnorm(length(xs),0,.15)  # add some noise
points(xy$x, xy$y, pch=19)                                        # observational data
                   
mo <- ModelObject(
        name = 'our eg',
        expr = expr
      )

# initial values for the parameters (we are assuming that we don't know the true values)
ps <- c(y0=0.05, u=1, tl=3, K=1, b0=0.1, b1=1, b2=0.01, k=0.5)

fit <- nlminb(start     = ps, 
              objective = function(p, data){
                            r = data$y - mo$value(p,data)
                            return(r %*% r)
                          }, 
              gradient  = mo$gradient, 
              hessian   = mo$hessian, 
              data      = xy)

lines(xy$x, mo$value(fit$par, xy), col="red", lwd=2)             # model estimate
```

And another eg:

```{r}
# make some data:
xy <- list(x=seq(0,10,by=0.25), y=dnorm(seq(0,10,by=0.25),10,2)) 
p0 <- c(y0=0.01, u=0.2, l=5, A=log(1.5/0.01))

mo <- ModelObject(
        name = 'gompertz',
        expr = expression( y0*exp(A*exp(-exp((u*exp(1)/A)*(l-x)+1))) )
      )

fit <- nlminb(start=p0, 
              objective= function(p, data){
                           r = data$y - mo$value(p,data)
                           return(r %*% r)
                         }, 
              gradient = mo$gradient, 
              hessian = mo$hessian, 
              data=xy)

plot(xy, main='Fit Results', pch=19);
lines(xy$x, mo$value(fit$par, xy), col="red", lwd=2)
```

