---
title: "Package magrittr"
author: "João Neto"
date: "December 2014"
output: html_document
---

Ref: [vignette](http://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html)

```{r, message=FALSE, warning=FALSE}
library(magrittr)
```


```{r}
car_data <- 
  mtcars %>%
  subset(hp > 100) %>%                                             # extract a subset
  aggregate(. ~ cyl, data = ., FUN = . %>% mean %>% round(2)) %>%  # aggregate info on n.cylinders
  transform(kpl = mpg %>% multiply_by(0.4251)) %>%                 # add new column
  print                                                            # print result before assign
```

Some notes:

+ By default the left-hand side (LHS) will be piped in as the first argument of the function appearing on the right-hand side (RHS). When the LHS is needed at a position other than the first, one can use the dot, `.`, as placeholder. This is used in the `aggregate` expression.

+ Whenever only one argument is needed, the LHS, then one can omit the empty parentheses (just like in `print`).

+ A pipeline with a dot (.) as LHS will create a unary function. This is used to define the aggregator function.

Other `.` egs:

```{r}
1:9 %>% paste(letters[.])
1:9 %>% paste(letters[.], .)
1:9 %>% { paste(letters[.]) } # use {} to prevent placing '.' as the 1st function argument
```

It's possible to use `%>%` into anonymous functions:

```{r}
car_data %>%
  { 
    if (nrow(.) > 0)
      rbind(head(., 1), tail(., 1))
    else .
  } %>%
  .[,1:3]
```

Whenever you want to use a function- or call-generating statement as right-hand side, parentheses are used to evaluate the right-hand side before piping takes place:

```{r}
1:10 %>% (substitute(f(), list(f = sum))) # with the outside (), it would be an error
```

## %T%

The "tee" operator, %T>% works like %>%, except it returns the left-hand side value, and not the result of the right-hand side operation. This is useful when a step in a pipeline is used for its side-effect (printing, plotting, logging, etc.).

```{r}
rnorm(200) %>%
  matrix(ncol = 2) %T>%
  plot %>%               # plot usually does not return anything. 
  colSums
```

## %$%

The "exposition" pipe operator, %$% exposes the names within the left-hand side object to the right-hand side expression. Essentially, it is a short-hand for using the with functions (and the same left-hand side objects are accepted).

```{r}
iris %>%
  subset(Sepal.Length > mean(Sepal.Length)) %$%
  cor(Sepal.Length, Sepal.Width)

data.frame(z = rnorm(100)) %$% 
  ts.plot(z)
```

## %<>%

The compound assignment pipe operator `%<>%` can be used as the first pipe in a chain. The effect will be that the result of the pipeline is assigned to the left-hand side object, rather than returning the result as usual. It is essentially shorthand notation for expressions like `foo <- foo %>% bar %>% baz`, which boils down to `foo %<>% bar %>% baz`.

```{r}
x <- 1:9
x %<>% sqrt %>% add(1)
x
```

Function `add(1)` is one of several aliases the package has to help writing expressions. Check the helpfile for more egs.
