---
title: "Memoization"
author: "João Neto"
date: "January 2015"
output: html_document
---

Memoization is the ability to cache the results of previous function invocations in order
to save time and space resources.

The classical eg is the naïve recursive computation of the Fibonacci sequence:

```{r}
# pre: n>=0
fib <- function(n) {
  if (n<2) 
    return(1)
  
  return (fib(n-1)+fib(n-2))
}

system.time(fib(25))
system.time(fib(30))
```

The problem is that the same arguments are computed again and again. If we were able to 
keep the intermediate results, the computation would be much faster:

```{r}
n <- 101
results <- rep(NA,n)  # intermediate results

fib2 <- function(n, results) {
  if (!is.na(results[n+1]))  # answer already known
    return (results[n+1])
  
  if (n<2) {
    eval.parent(substitute(results[n+1] <- 1))  # needed: R does not have call by reference
  } else {
    eval.parent(substitute(results[n+1] <- fib2(n-1, results) + fib2(n-2, results)))
  }
  
  return (results[n+1])
}

system.time(fib2(25, results))
system.time(fib2(30, results))
system.time(fib2(100, results))
```

There is a R package useful to memoize functions:

```{r}
library(memoise)
```

There are just three functions:

+ memoise -- memoise a function

+ forget -- resets the cache of a memoised function

+ is.memoised -- checks if a function is memoised

```{r}
a <- function(n) { runif(n) }

memA <- memoise(a)

replicate(5,    a(2))
replicate(5, memA(2))
```

Notice, however, that it does not work that well with recursivity:

```{r}
fibM <- memoize(fib)

system.time(fibM(25))
system.time(fibM(30))
system.time(fibM(33))
```

In this [post](http://adamleerich.com/2014/12/07/fibonacci-sequence-in-r-with-memoization/) there's an alternative solution:

```{r}
fibM <- (function() {
 
    # The code here related to the cache *mostly* comes from the memoise
    # package's object new_cache.
 
    cache <- NULL
 
    cache_reset <- function() {
        cache <<- new.env(TRUE, emptyenv())
        cache_set('0', 0)
        cache_set('1', 1)
    }
 
    cache_set <- function(key, value) {
        assign(key, value, envir = cache)
    }
 
    cache_get <- function(key) {
        get(key, envir = cache, inherits = FALSE)
    }
 
    cache_has_key <- function(key) {
        exists(key, envir = cache, inherits = FALSE)
    }
 
    cache_reset() # Initialize the cache
  
    # This is the function that gets returned by the anonymous function and
    # becomes fibM.
    function(n) {
 
        nc <- as.character(n)
 
        # Handle "vectors" by element
        if (length(n) > 1) {
            return(sapply(n, fibM))
        }
 
        # Cached cases
        if (cache_has_key(nc)) 
            return(cache_get(nc))
 
        out <- fibM(n - 1) + fibM(n - 2)
        cache_set(nc, out)
        return(out)
    }
})()
```

Let's use it:

```{r}
ls(environment(fibM)$cache) # current environment (only base values are computed)
fibM(30)
ls(environment(fibM)$cache)

system.time(fibM(33))
```

