---
title: "Relational Algebra in R"
author: "João Neto"
date: October 2014
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 6
cache: yes
---


[wikipedia] _In computer science, relational algebra is an offshoot of first-order logic and of algebra of sets concerned with operations over finitary relations, usually made more convenient to work with by identifying the components of a tuple by a name (called attribute) rather than by a numeric column index, which is what is called a relation in database terminology.

The main application of relational algebra is providing a theoretical foundation for relational databases, particularly query languages for such databases, chiefly among which is SQL._

Package `sqldf`
-------------

This package performs SQL queries on R data frames

```{r, tidy=FALSE}
library(sqldf)
set.seed(101)
size = 50
get.name <- function(size)paste(sample(letters,size),collapse="")
mydf <- data.frame(id=1:size,
                   name=as.vector(Map(get.name, rep(5,size)), mode="character"),
                   time=rpois(size,100),
                   size=floor(runif(size,1,6)))
head(mydf)
sqldf("SELECT * 
       FROM mydf 
       LIMIT 6")
sqldf("SELECT * 
       FROM mydf 
       WHERE name IN ('megku','qdaym')")
sqldf("SELECT * 
       FROM mydf 
       WHERE name LIKE '%a%'")  # names with at least one 'a'
sqldf("SELECT id, MAX(time) 
       FROM mydf")
sqldf("SELECT * 
       FROM mydf 
       WHERE time BETWEEN 90 AND 93")
sqldf("SELECT * 
       FROM mydf 
       WHERE name BETWEEN 'a' AND 'g'")
sqldf("SELECT time, count(*) AS n_refs
       FROM mydf 
       GROUP BY time HAVING count(*) > 2")
sqldf("SELECT * 
       FROM mydf 
       WHERE time = 103")
sqldf("SELECT id, time
       FROM mydf 
       WHERE time >= 70 AND time <= 86 
       ORDER BY time;")
sqldf("SELECT size, avg(time) as avg_time 
       FROM mydf 
       GROUP BY size;")

size2 <- 200
mydf2 <- data.frame(id=sample(1:50,size2,replace=TRUE),
                    value=rexp(size2,1/50))
head(mydf2)
sqldf("SELECT count(*) AS n_values
       FROM mydf2 
       GROUP BY id
       LIMIT 12")
sqldf("SELECT id, SUM(value) AS sum_vals
       FROM mydf2 
       GROUP BY id
       HAVING SUM(value) > 300")

# Let the joins begin!
# INNER JOIN, ie, returns rows when there is at least one match in both tables
sqldf("SELECT mydf.name, mydf2.value 
       FROM mydf JOIN mydf2 
       ON mydf.id = mydf2.id 
       ORDER BY mydf.name 
       LIMIT 12")

# LEFT (OUTER) JOIN keyword returns all rows from the left table (table_name1), 
# even if there are no matches in the right table (table_name2).
sqldf("SELECT mydf.name, mydf2.value 
       FROM mydf LEFT JOIN mydf2 
       ON mydf.id = mydf2.id 
       ORDER BY mydf.name 
       LIMIT 12")  # in this case there is no difference, all id's have values in mydf2

mydf3 <- data.frame(id=sample(51:60,size2,replace=TRUE),
                    value=rexp(size2,1/50))

# UNION combines the result-set of two or more SELECT statements
# It does not include repetitions (for that, use UNION ALL)
sqldf("SELECT mydf2.id FROM mydf2 
       UNION
       SELECT mydf3.id FROM mydf3
       ORDER BY id DESC
       LIMIT 12") 
```

### Reading CSV files

```{r, warning=FALSE}
df <- read.csv("effort.csv")
head(df, 10)
df <- read.csv.sql("effort.csv", "SELECT * FROM file WHERE effort>10")
head(df, 10)
df <- read.csv.sql("effort.csv", "SELECT X, effort FROM file")
head(df, 10)
df <- read.csv.sql("effort.csv", "SELECT X, effort FROM file ORDER BY X DESC")
head(df, 10)
df <- read.csv.sql("effort.csv", "SELECT effort, COUNT(*) AS count FROM file GROUP BY effort")
head(df, 10)
```

Package `relations`
------------------

_Data Structures and Algorithms for for k-ary relations with arbitrary domains, featuring relational algebra, predicate functions, and fitters for consensus relations_

Check [http://cran.r-project.org/web/packages/relations/index.html](http://cran.r-project.org/web/packages/relations/index.html)

```{r, tidy=FALSE}
library(relations)
PersonDF <-
  data.frame(Name = c("Harry", "Sally", "George", "Helena", "Peter"),
             Age = c(34, 28, 29, 54, 34),
             Weight = c(80, 64, 70, 54, 80),
             stringsAsFactors = FALSE)
Person <- as.relation(PersonDF)
## see relation
relation_table(Person)
## projection
relation_table(relation_projection(Person, c("Age", "Weight")))
## selection
relation_table(R1 <- relation_selection(Person, Age < 29))
relation_table(R2 <- relation_selection(Person, Age >= 34))
relation_table(R3 <- relation_selection(Person, Age == Weight))
## union
relation_table(R1 %U% R2)
## works only for the same domains:
relation_table(R2 | R3)
## complement
relation_table(Person - R2)
## intersection
relation_table(relation_intersection(R2, R3))
## works only for the same domains:
relation_table(R2 & R3)
## symmetric difference
relation_table(relation_symdiff(R2, R3))
## cartesian product
Employee <-
  data.frame(Name = c("Harry", "Sally", "George", "Harriet", "John"),
             EmpId = c(3415, 2241, 3401, 2202, 3999),
             DeptName = c("Finance", "Sales", "Finance", "Sales", "N.N."),
             stringsAsFactors = FALSE)

Employee <- as.relation(Employee)
relation_table(Employee)
Dept <- data.frame(DeptName = c("Finance", "Sales", "Production"),
                   Manager = c("George", "Harriet", "Charles"),
                   stringsAsFactors = FALSE)
Dept <- as.relation(Dept)
relation_table(Dept)
relation_table(Employee %><% Dept)
## Natural join
relation_table(Employee %|><|% Dept)
## left (outer) join
relation_table(Employee %=><% Dept)
## right (outer) join
relation_table(Employee %><=% Dept)
## full outer join
relation_table(Employee %=><=% Dept)
## antijoin
relation_table(Employee %|>% Dept)
relation_table(Employee %<|% Dept)
## semijoin
relation_table(Employee %|><% Dept)
relation_table(Employee %><|% Dept)
## division
Completed <-
  data.frame(Student = c("Fred", "Fred", "Fred", "Eugene",
                         "Eugene", "Sara", "Sara"),
             Task = c("Database1", "Database2", "Compiler1",
                      "Database1", "Compiler1", "Database1",
                      "Database2"),
             stringsAsFactors = FALSE)
Completed <- as.relation(Completed)
relation_table(Completed)
DBProject <- data.frame(Task = c("Database1", "Database2"),
                        stringsAsFactors = FALSE)
DBProject <- as.relation(DBProject)
relation_table(DBProject)
relation_table(Completed %/% DBProject)
## division remainder
relation_table(Completed %% DBProject)
```



