---
title: "Domain-Specific Languages"
author: "João Neto"
date: "December 2016"
output: html_document
---

From Wikipedia:

> A domain-specific language (DSL) is a computer language specialized to a particular application domain. This is in contrast to a general-purpose language (GPL), which is broadly applicable across domains. 

R's flexibility makes it possible to create DSL's inside the language. Details can be found at Hadley Wickham's [Advanced R](http://adv-r.had.co.nz/dsl.html).

Here we make some egs.

Wires
=====

Refs: [Retaining a Sense of Wonder](http://www.cs.uni.edu/~wallingf/blog/archives/monthly/2016-12.html#e2016-12-21T14_31_14.htm)

In [Advent of Code](http://adventofcode.com/) there's a [problem](http://adventofcode.com/2015/day/7) about finding the values produced by a network of logic gates that are described like this:

```{r}
problem <-
"   x AND y -> d       
    x OR y -> e        
    x LSHIFT 2 -> f    
    y RSHIFT 2 -> g    
    NOT x -> h         
    NOT y -> i         
    123 -> x           
    456 -> y     
"
```

there are the typical logic gates and we wish to find the values for each gate, after all are executed.

One way would be to create a graph of the described gates, make a topological sort to find what is the order of execution, before executing them, one by one. This means some code (we could use the `igraph` package to help).

However, there's a more elegant solution: _turn the data into a program_. The trick is to transform each line into a function! That way, if we wish to find what is the value of `123 -> x` we just execute function `x() <- function 123`. The order which they are executed becomes irrelevant, each function call decides which are the needed functions, since _functions defer computation_.

This is something easy to do in LISP, but it turns out that R also makes it easy with its meta-programming tools. 

So, let's first define the operations required by the problem:

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(stringr)

int2unint <- function(n) ifelse(n>=0,n,65536+n)   # the DSL uses unsigned 16-bit ints

# the basic commands for the DSL
not    <- purrr::compose(int2unint,bitwNot)       # using R bitwise operators
and    <- purrr::compose(int2unint,bitwAnd)       
or     <- purrr::compose(int2unint,bitwOr)
lshift <- purrr::compose(int2unint,bitwShiftL)
rshift <- purrr::compose(int2unint,bitwShiftR)
```

Next we need functions that convert each line into a R function description:

```{r}
# Each command has an appropriate description builder that converts 
# the token list into a string with a R function definition, eg: 
#   > tokens
#   [1] "456" "->"  "y"
#   > create.const.description(tokens)
#   [1] "y <- function() 456"

create.and.description <- function(tokens)
  paste0(tokens[5], " <- function() and(",    tokens[1], "(),", tokens[3], "())")

create.or.description <- function(tokens)
  paste0(tokens[5], " <- function() or(",     tokens[1], "(),", tokens[3], "())")

create.lshift.description <- function(tokens)
  paste0(tokens[5], " <- function() lshift(", tokens[1], "(),", tokens[3], ")")

create.rshift.description <- function(tokens)
  paste0(tokens[5], " <- function() rshift(", tokens[1], "(),", tokens[3], ")")

create.not.description <- function(tokens)
  paste0(tokens[4], " <- function() not(",    tokens[2], "())")

create.const.description <- function(tokens)
  paste0(tokens[3], " <- function() ", tokens[1])
```

An example:

```{r}
"y RSHIFT 2 -> g" %>% 
  str_split(" ") %>% 
  `[[`(1) %>% 
  create.rshift.description()
```

Notice that until now this R function description is just a string, we need to convert it to an expression so we can evaluate it. This is done by the next function:

```{r}
# given a list of tokens, create the respective R function in the global enviroment
create.fun <- function(tokens) {
  
       if(any(str_detect(tokens,"AND"   ))) op <- create.and.description
  else if(any(str_detect(tokens,"OR"    ))) op <- create.or.description
  else if(any(str_detect(tokens,"LSHIFT"))) op <- create.lshift.description
  else if(any(str_detect(tokens,"RSHIFT"))) op <- create.rshift.description
  else if(any(str_detect(tokens,"NOT"   ))) op <- create.not.description
  else                                      op <- create.const.description
  
  # evals executes this new R command, creating the function
  eval(parse(text=op(tokens)), envir=globalenv())  
}
```

Let's place the parsing and evaluing into a function:

```{r}
exec_gates <- function(problem) {
  # read data and tokenize it, line by line
  read_lines(problem) %>% 
    purrr::map(function(line) line %>% str_trim %>% str_split("[ \t]+")) -> list.tokens 
  
  results <- data.frame(var=rep(NA,length(list.tokens)))
  
  for(.i in 1:length(list.tokens)) {
    tokens <- list.tokens[[.i]][[1]]            # get tokens from i-th line
    results[.i,1] <- tokens[length(tokens)]     # save variable name into results df
    create.fun(tokens)                          # create function
  }
  
  results$var %>%                               # execute functions and save their results
    purrr::map(function(var) eval(parse(text=paste0(var,"()")))) -> results$val
  results
}
```

And now we are ready to execute the problem:

```{r}
exec_gates(problem)
```










