---
title: "Connecting with Javascript"
author: "João Neto"
date: September 2015
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 6
---

Refs: 

+ https://cran.r-project.org/web/packages/V8/vignettes/v8_intro.html

Library `v8` is a Google's open source JavaScript engine written in C++. The R package `V8` provides a standalone JavaScript for R.

```{r}
library(V8)
```

To execute JavaScript code we need a handler for a context, which is an execution environment:

```{r}
ct <- new_context();
```

Within it, we can start programming:

```{r}
ct$eval("var a = 1")
ct$get("a")

ct$call("function(x, y){return x * y}", 123, 3)
```

Make functions:

```{r}
# make factorial function
# JS("...") means JavaScript literal code
ct$assign("fact",JS("
  function(n) {
    if (n==0)
      return 1;
    else
      return n*fact(n-1);
  }
"))

# use it
ct$eval("fact(6)")
```

We can export R data to JavaScript:

```{r, collapse=TRUE}
head(cars)
ct$assign("mydata", cars)
head(ct$get("mydata"))
```

Validate expressions without evaluation:

```{r, collapse=TRUE}
ct$validate("function foo(x){2x}")
ct$validate("function foo(x){2*x}")
ct$validate("function(x){2*x}")
ct$validate("(function(x){2*x})") # for anonymous functions, wrap them in ()'s
```

We can create JSON descriptions of some data structure:

```{r}
ct$eval("JSON.stringify({x:Math.random()})")
```

Use JavaScript libraries:

```{r}
ct$source(system.file("js/underscore.js", package="V8")) # js lib for functional programming
ct$call("_.filter", mtcars, JS("function(x){return x.mpg < 15}"))
```

Place new js libraries under `[your R]\library\V8\js`.

Notice that we can have more than one context:

```{r, collapse=TRUE}
ct$eval("var x = 1")

ct2 <- new_context();
ct2$eval("var x = 2") # the variable is in a 2nd context

ct$get("x")
ct2$get("x")
```

To erase a context use `ct$reset()`.

It's also possible to make an interactive console by typing `ct$console()`. To end it, type `exit`.

Another eg where we create and use a composition operator for tuples (this is a pseudo-monad in JavaScript):

```{r, collapse=TRUE}
ct$eval('
     // tuple composition
     gName = function(name) {   // change name
          return function(g) { 
              g.name = name; 
              return g; 
          }; 
     };
     
     gType = function(type) {   // change type
          return function(g) { 
              g.type = type; 
              return g; 
          }; 
     };
      
     c = function(g, op) {      // tuple composition operator
          return op(g)
     };

     // show tuple

     showtuple = function(g) {
          return "tuple: {name:" + g.name + ", type:" + g.type + "}";
     };
              
     // make a tuple 

     tuple = { name : "a",
               type : "b" };  // a tuple eg
')

ct$eval('showtuple(tuple);')
ct$eval('newTuple = c(c(tuple, gName("new_a")), gType("new_b"));')
ct$eval('showtuple(newTuple);')
```

