---
title: "Connecting with Python"
author: "João Neto"
date: September 2015
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 6
cache: yes
---


Refs:

+ [Calling Python from R using rPython](http://www.programmingr.com/content/calling-python-r-rpython/)

+ (http://rpython.r-forge.r-project.org/)[http://rpython.r-forge.r-project.org/]

The `rPython` package allows calling Python from R. THe Python session that the package iniciates lives during the current R session.

The installation on Windows requires extra steps and they are explained [here](https://github.com/cjgb/rPython-win). The next code snippet was part of this process and only needs to be done once:

```{r, eval=FALSE}
# install Rtools (if necessary)
install.packages("installr")
library(installr)
install.Rtools()

# install devtools (if necessary)
install.packages("devtools")
library(devtools)

# after download of the package & config/folder updates
# (check https://github.com/cjgb/rPython-win), execute:
install("C:/Users/jpn.INFORMATICA/Software/_Langs/rPython")
```

After that, we just need to import the package, as usual:

```{r}
library(rPython)
```

An example of function definition and call:

```{r}
python.exec("
  def fact(n): 
    if n==0:
      return 1
    else:
      return n*fact(n-1)
")

python.call("fact", 6)
```

Lists are translated automatically:

```{r}
python.call( "len", 1:3 )
a <- 1:4
b <- 5:8
python.exec( "def concat(a,b): return a+b" )
python.call( "concat", a, b)
```

We can assign values to python vars, and use them in method calls:

```{r}
str <- "hello world"
python.assign( "a",  str )
python.method.call( "a", "split", " " )
```

Python files can be loaded using `python.load`.

We can also make imports:

```{r}
python.exec( "import math" )
my_pi <- python.get( "math.pi" )
my_pi
```

A stats eg in which Python does a linear regression:

```{r}
head(cars)
plot(cars, pch=20)

python.assign( "X",  cars$speed )
python.assign( "Y",  cars$dist )

python.exec("
    import numpy as np
    from sklearn import linear_model

    X = np.reshape(X, (50,1))  # reshape to 50x1 vector
    Y = np.reshape(Y, (50,1))

    regr = linear_model.LinearRegression()
    regr.fit(X, Y)

    intcp = regr.intercept_[0]
    coefs = regr.coef_[0][0]
")

intcp <- python.get( "intcp " )
coefs <- python.get( "coefs" )

abline(intcp, coefs, col="red", lwd=2)
```

