---
title: "Symbolic Computation in R"
author: "João Neto"
date: October 2014
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 6
cache: yes
---

Herein we use package `rSymPy` that needs Python and Java instalattion (this library uses SymPy via Jython). If necessary cf. [troubleshooting](http://stackoverflow.com/questions/2399027/cannot-load-rjava-because-cannot-load-a-shared-library).

Refs: 

+ Reference manual (very short) [http://cran.r-project.org/web/packages/rSymPy/rSymPy.pdf](http://cran.r-project.org/web/packages/rSymPy/rSymPy.pdf)
+ SymPy Tutorial: [http://docs.sympy.org/latest/tutorial/](http://docs.sympy.org/latest/tutorial/)

```{r, warning=FALSE, message=FALSE}
library(rSymPy)  
```

```{r}
x <- Var("x")
x+x
x*x/x
y <- Var("x**3")
x/y
z <- sympy("2.5*x**2")
z + y
sympy("sqrt(8).evalf()")  # evaluate an expression
sympy("sqrt(8).evalf(50)")
sympy("pi.evalf(120)")
sympy("one = cos(1)**2 + sin(1)**2")
sympy("(one - 1).evalf()")  # rounding errors
sympy("(one - 1).evalf(chop=True)")  # rouding this type of roundoff errors
sympy("Eq(x**2+2*x+1,(x+1)**2)") # create an equation
sympy("a = x**2+2*x+1")
sympy("b = (x+1)**2")
"0" == sympy("simplify(a-b)")  # if they are equal, the result is zero
# simplify works in other tasks:
sympy("simplify((x**3 + x**2 - x - 1)/(x**2 + 2*x + 1))")
sympy("(x + 2)*(x - 3)")
sympy("expand((x + 2)*(x - 3))")
sympy("factor(x**3 - x**2 + x - 1)")
y <- Var("y")
z <- Var("z")
sympy("collect(x*y + x - 3 + 2*x**2 - z*x**2 + x**3, x)")  # organize equation around var 'x'
sympy("(x*y**2 - 2*x*y*z + x*z**2 + y**2 - 2*y*z + z**2)/(x**2 - 1)")
sympy("cancel((x*y**2 - 2*x*y*z + x*z**2 + y**2 - 2*y*z + z**2)/(x**2 - 1))")
sympy("expand_func(gamma(x + 3))")
sympy("y = x*x") # create a variable 'y' in the SymPy persistant state
sympy("A = Matrix([[1,x], [y,1]])")
sympy("A**2")
sympy("B = A.subs(x,1.1)")  # replace x by 1.1 (btw, SymPy objects are immutable)
sympy("B**2")
# more replacement, a subexpression by another:
sympy("expr = sin(2*x) + cos(2*x)")
sympy("expr.subs(sin(2*x), 2*sin(x)*cos(x))")
sympy("expr.subs(x,pi/2)")
# more matrix stuff:
a1 <- Var("a1")
a2 <- Var("a2")
a3 <- Var("a3")
a4 <- Var("a4")

A <- Matrix(List(a1, a2), List(a3, a4))
#define inverse and determinant
Inv <- function(x) Sym("(", x, ").inv()")
Det <- function(x) Sym("(", x, ").det()")

A
cat(A,"\n")
Inv(A)
Det(A)
# create function exponential
Exp <- function(x) Sym("exp(", x, ")") 
Exp(-x) * Exp(x) 

y <- Var("y")
sympy("sqrt(8)")             # simplify expression
sympy("solve(x**2 - 2, x)")  # solve x^2-2=0
sympy("limit(1/x, x, oo)")   # limit eg, x -> Inf
sympy("limit(1/x, x, 0)")  
sympy("integrate(exp(-x))")              # indefinite integral
sympy("integrate(exp(-x*y),x)")          # indefinite integral
sympy("integrate(exp(-x), (x, 0, oo))")  # definite integral
integrate( function(x) exp(-x), 0, Inf)  # integration is possible in R
sympy("integrate(x**2 - y, (x, -5, 5), (y, -pi, pi))")  # definite integral
sympy("diff(sin(2*x), x, 1)")  # first derivative
D( expression(sin(2*x)), "x" ) # also possible in base R
sympy("diff(sin(2*x), x, 2)")  # second derivative
sympy("diff(sin(2*x), x, 3)")  # third  derivative
sympy("diff(exp(x*y*z), x, y, y)") # d^3/dxdy^2
sympy("diff(exp(x*y*z), x, z, 3)") # d^4/dxdz^3
sympy("(1/cos(x)).series(x, 0, 10)")  # taylor expansion
sympy("exp(x).series(x, 0, 5)")       # taylor expansion
sympy("exp(x).series(x, 0, 5).removeO()")
sympy("Matrix([[1, 2], [2, 2]]).eigenvals()")  # get eigenvalues of matrix
sympy("latex(Integral(cos(x)**2, (x, 0, pi)))")
```

This can be used within the markup text 
`r sympy("latex(Integral(cos(x)**2,  (x, 0, pi)))")` = 
`r sympy("latex(integrate(cos(x)**2, (x, 0, pi)))")`


## Recognizing numbers:

`nsimplify` takes a floating point number and tries to simplify it:

```{r}
sympy("nsimplify(4.242640687119286)")
sympy("nsimplify(cos(pi/6))")
sympy("nsimplify(6.28, [pi], tolerance=0.01)")
sympy("nsimplify(pi, tolerance=1e-5)")
sympy("nsimplify(pi, tolerance=1e-6)")
sympy("nsimplify(29.60881, constants=[pi,E], tolerance=1e-5)")
```

## Euler's formula

As seen above, `sympy` allows for the Taylor's expansion.

To check on Euler's formula

$$e^{i\theta} = cos~\theta + i ~ sin~\theta$$

let's define Taylor series for $sin, cos, e^{i\theta}$,

```{r}
theta <- Var("theta")

sin.series <- function(n) sympy(paste0(  "sin(theta).series(theta, 0, ", n, ")"))
cos.series <- function(n) sympy(paste0(  "cos(theta).series(theta, 0, ", n, ")"))
exp.series <- function(n) sympy(paste0("exp(I*theta).series(theta, 0, ", n, ")"))
```

We can see similarities from these expansions:

$sin~\theta$ = `r sympy(paste0("latex(",sin.series(8),")"))`

$cos~\theta$ = `r sympy(paste0("latex(",cos.series(8),")"))`

$e^{i\theta}$ = `r sympy(paste0("latex(",exp.series(8),")"))`

let's separate the terms according to $i$,

$e^{i\theta}$ = `r sympy(paste0("latex(collect(",exp.series(8),",I))"))`

and we see that it seems to follow

$$e^{i\theta} = cos~\theta + i ~ sin~\theta$$

We can check it by subtracting both expressions and confirm that all terms cancel:

```{r}
n <- 50

# expansion(sin)*i
isin.series <- 
  function(n) sympy(paste0("expand(I*sin(theta).series(theta, 0, ", n, "))"))

sympy( paste0( "nsimplify(",
               exp.series(n),
               "-(",
               cos.series(n),"+", isin.series(n),
               "))"   
              ) 
     )
```

`r sympy( paste0( "latex(nsimplify(",
               exp.series(n),
               "-(",
               cos.series(n),
               "+ ", isin.series(n),
               ")))"   
              ) 
     )`