---
title: "Finding a change point"
author: "João Neto"
date: "September 2016"
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 6
cache: yes
---

Suppose there is a process that is modeled by two linear regressions where exists a change point between the two regimes. 

Here's an example:

```{r}
# Construct sample data
set.seed(121)
a1 <- 1
a2 <- 1.5
b1 <- 0
b2 <- -.15

n <- 101
changepoint <- 30

x  <- seq(0,1,len=n)
y1 <- a1 * x[1:changepoint]     + b1
y2 <- a2 * x[(changepoint+1):n] + b2
y  <- c(y1,y2) + rnorm(n,0,.05)

plot(x, y, pch=19)
abline(a=b1, b=a1, lwd=2, col="orange")
abline(a=b2, b=a2, lwd=2, col="blue")
abline(v=x[changepoint], lty=2)
```

We wish to find this change point using probabilistic programming. Herein, we'll use BUGS.

```{r, message=FALSE}
library(BRugs)

run.model <- function(model, samples, data=list(), chainLength=10000, burnin=0.10, 
                      init.func, n.chains=1, thin=1) {
  
  writeLines(model, con="model.txt")  # Write the modelString to a file
  modelCheck( "model.txt" )           # Send the model to BUGS, which checks the model syntax
  if (length(data)>0)                 # If there's any data available...
    modelData(bugsData(data))         # ... BRugs puts it into a file and ships it to BUGS
  modelCompile(n.chains)              # BRugs command tells BUGS to compile the model
  
  if (missing(init.func)) {
    modelGenInits()                   # BRugs command tells BUGS to randomly initialize a chain
  } else {
    for (chain in 1:n.chains) {       # otherwise use user's init data
      modelInits(bugsInits(init.func))
    }
  }
  
  modelUpdate(chainLength*burnin)     # Burn-in period to be discarded
  samplesSet(samples)                 # BRugs tells BUGS to keep a record of the sampled values
  samplesSetThin(thin)                # Set thinning
  modelUpdate(chainLength)            # BRugs command tells BUGS to randomly initialize a chain
}

```

## The Model

The model states that there is one change point, $\tau$, that separates the two linear regimes. 

Let's assume that both regimes share the same variance (which might not be the case):

$$y_i \sim \mathcal{N}(a_1 x_i + b_1, \sigma^2), ~ ~ i \leq \tau$$
$$y_i \sim \mathcal{N}(a_2 x_i + b_2, \sigma^2), ~ ~ i \gt  \tau$$

$$a_i \sim \mathcal{N}(\alpha_a, \beta_a), ~ i=1,2$$
$$b_i \sim \mathcal{N}(\alpha_b, \beta_b), ~ i=1,2$$

$$\tau \sim \text{DiscreteUniform}(x_{init}, x_{end})$$

Notice that parameters $a_1$ and $a_2$ are assumed to belong to similar distributions. This is reasonable, since if that was not the case, it would be easy to identify the changepoint and all this would not be needed.

I will not define $\alpha, \beta$ as hyperparameters, but will just choose reasonable values looking at the available data.

This model can be written in Bugs like this:

```{r, collapse=TRUE}
modelString = "
    model {
      tau ~ dcat(xs[])        # the changepoint
      
      a1 ~ dnorm(1,3)
      a2 ~ dnorm(1,3)
      b1 ~ dnorm(0,2)
      b2 ~ dnorm(0,2)
  
      for(i in 1:N) {
        xs[i]  <- 1/N    # all x_i have equal priori probability to be the changepoint
  
        mu[i]  <- step(tau-i) * (a1*x[i] + b1) + 
                  step(i-tau-1) * (a2*x[i] + b2)

        phi[i] <- -log( 1/sqrt(2*pi*sigma2) * exp(-0.5*pow(y[i]-mu[i],2)/sigma2) ) + C

        dummy[i] <- 0
        dummy[i] ~ dpois( phi[i] )
      }
  
      sigma2 ~ dunif(0.001, 2)
      C  <- 100000
      pi <- 3.1416
    }
"

data.list = list(
  N = length(x),
  x = x,
  y = y
)

run.model(modelString, samples=c("tau", "a1", "a2", "b1", "b2", "sigma2"), 
          data=data.list, chainLength=5e4, burnin=0.2, n.chains=4, thin=2)
```

After running it, let's extract the mcmc chains and get their summaries:

```{r}
stats <- samplesStats(c("tau", "a1", "a2", "b1", "b2", "sigma2"))
stats

tau.hat <- stats[1,5]
a1.hat  <- stats[2,5]
a2.hat  <- stats[3,5]
b1.hat  <- stats[4,5]
b2.hat  <- stats[5,5]

plot(x, y, pch=19)
abline(a=b1,     b=a1,     lwd=1, lty=2, col="orange")
abline(a=b2,     b=a2,     lwd=1, lty=2, col="blue")
abline(a=b1.hat, b=a1.hat, lwd=2,        col="orange")
abline(a=b2.hat, b=a2.hat, lwd=2,        col="blue")
abline(v=x[changepoint], lty=2, col="grey")
abline(v=x[tau.hat],     lty=2)
```

We can also use some tools to check if the produced chains are reasonable:

```{r}
library(coda)

samplesHistory("tau", mfrow = c(1, 1))
samplesAutoC("tau", mfrow = c(1, 1), 1)
chain <- samplesSample("tau")
hist(chain[2e4:10e4], breaks=40, prob=T)
```

Notice that there's some probabilistic mass over 100 and a bit over zero. This is not unexpected since, for this data, the transition at the changepoint is smooth enough to consider the hypothesis that there's no changepoint.

## Using extra information

If we know that both linear regimes meet at $x[\tau]$ we can include that information in the model. Also, here I'll replace $a_2$ with a delta over $a_1$:

$$y_i \sim \mathcal{N}(a_1 x_i + b_1, \sigma^2), ~ ~ i \leq \tau$$
$$y_i \sim \mathcal{N}((a_1+\delta) x_i + (b_1 - \delta x[\tau]), \sigma^2), ~ ~ i \gt  \tau$$

$$a_1 \sim \mathcal{N}(\alpha_a, \beta_a)$$
$$b_1 \sim \mathcal{N}(\alpha_b, \beta_b)$$

$$\delta \sim \mathcal{N}(0, 2),$$

$$\tau \sim \text{DiscreteUniform}(x_{init}, x_{end})$$

```{r, collapse=TRUE}
modelString = "
    model {
      tau ~ dcat(xs[])        # the changepoint
      
      a1 ~ dnorm(1,3)
      b1 ~ dnorm(0,2)

      delta ~ dnorm(0,1)

      for(i in 1:N) {
        xs[i]  <- 1/N    # all x_i have equal priori probability to be the changepoint
  
        mu[i]  <- step(tau-i)   * (  a1      *x[i] +     b1       ) + 
                  step(i-tau-1) * ((a1+delta)*x[i] + (b1 - delta*x[tau]))

        phi[i] <- -log( 1/sqrt(2*pi*sigma2) * exp(-0.5*pow(y[i]-mu[i],2)/sigma2) ) + C

        dummy[i] <- 0
        dummy[i] ~ dpois( phi[i] )
      }
  
      sigma2 ~ dunif(0.001, 2)
      C  <- 100000
      pi <- 3.1416
    }
"

data.list = list(
  N = length(x),
  x = x,
  y = y
)

run.model(modelString, samples=c("tau", "a1", "b1", "delta", "sigma2"), 
          data=data.list, chainLength=5e4, burnin=0.2, n.chains=1, thin=2)
```

```{r}
stats <- samplesStats(c("tau", "a1", "b1", "delta", "sigma2"))
stats

tau.hat    <- stats[1,5]
a1.hat     <- stats[2,5]
b1.hat     <- stats[3,5]
delta.hat  <- stats[4,5]

plot(x, y, pch=19)
abline(a=b1, b=a1, lwd=1, lty=2, col="orange")
abline(a=b2, b=a2, lwd=1, lty=2, col="blue")

abline(a=b1,                      b=a1,           lwd=2, col="orange")
abline(a=b1-delta.hat*x[tau.hat], b=a1+delta.hat, lwd=2, col="blue")

abline(v=x[changepoint], lty=2, col="grey")
abline(v=x[tau.hat],     lty=2)
```

