---
title: "Mapping stuff"
date: "May, 2015"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

Refs:

+ [Maps in R: Introduction - Drawing the map of Europe](http://www.milanor.net/blog/?p=534)

+ [Maps in R: Plotting data points on a map](http://www.milanor.net/blog/?p=594)

+ [Maps in R: choropleth maps](http://www.milanor.net/blog/?p=634)

```{r}
# download from http://cran.r-project.org/web/packages/rworldmap/index.html
# and high def map from http://cran.r-project.org/web/packages/rworldxtra/index.html
library(rworldmap)
newmap <- getMap(resolution = "low")
plot(newmap)
```

## Defining map limits

```{r}
library(ggmap)

europe.limits <- geocode(c(
  "CapeFligely,RudolfIsland,Franz Josef Land,Russia",
  "Gavdos,Greece",
  "Faja Grande,Azores",
  "SevernyIsland,Novaya Zemlya,Russia")
)
europe.limits
```

```{r}
plot(newmap,
  xlim = range(europe.limits$lon),
  ylim = range(europe.limits$lat),
  asp = 1.0
)
```

## Adding information on map

Get some data, in this case a dataset including airport coordinates:

```{r}
# from http://openflights.org/data.html
airports <- read.csv("airports.dat", header = FALSE)
colnames(airports) <- c("ID", "name", "city", "country", "IATA_FAA", "ICAO", "lat", "lon", "altitude", "timezone", "DST")
head(airports)
```

Now place them in the map

```{r, fig.height=8, fig.width=12}
plot(newmap, xlim = range(europe.limits$lon), ylim = range(europe.limits$lat), asp = 1.0)
points(airports$lon, airports$lat, col = "red", cex = .25)
```

## Using ggmap

Get a map from Google Maps (there's also other servers)

```{r}
library(ggmap)
library(mapproj)
map <- get_map(location = 'Europe', zoom = 4)
ggmap(map)
```

Get arrival and departure info to attach to the airpot data:

```{r}
# Also from http://openflights.org/data.html
routes <- read.csv("routes.dat", header=F)
colnames(routes) <- c("airline", "airlineID", "sourceAirport", "sourceAirportID", "destinationAirport", "destinationAirportID", "codeshare", "stops", "equipment")
head(routes)
```

Compute how many arrivals and departures from each airport (the joining key is ID)

```{r}
library(plyr)
departures <- ddply(routes, .(sourceAirportID), "nrow")
names(departures)[2] <- "flights"
head(departures)
arrivals <- ddply(routes, .(destinationAirportID), "nrow")
names(arrivals)[2] <- "flights"
head(arrivals)
```

```{r}
airportD <- merge(airports, departures, by.x = "ID", by.y = "sourceAirportID")
head(airportD)
airportA <- merge(airports, arrivals,   by.x = "ID", by.y = "destinationAirportID")
```

A ggmap is an object of class ggplot, so typical ggplot operations work with it

```{r}
mapPoints <- ggmap(map) +
             geom_point(aes(x = lon, y = lat, size = sqrt(flights)), data = airportD, alpha = .5) +
             scale_size_area(breaks = sqrt(c(1, 5, 10, 50, 100, 500)), 
                             labels = c(1, 5, 10, 50, 100, 500), name = "departing routes")
mapPoints
```

## Using shapefiles with maptools

```{r}
library(maptools) # read shapefiles (.shp)
library(ggplot2)
library(ggmap)

# from http://ec.europa.eu/eurostat/web/gisco/geodata/reference-data/administrative-units-statistical-units
# read administrative boundaries (change folder appropriately)
eurMap <- readShapePoly(fn="NUTS_2010_60M_SH/data/NUTS_RG_60M_2010")
plot(eurMap)
# convert shp data to data frame for ggplot
eurMapDf <- fortify(eurMap, region='NUTS_ID')

# read downloaded data (change folder appropriately)
eurEdu <- read.csv("educ_thexp_1_Data.csv", stringsAsFactors = F)
eurEdu$Value <- as.double(eurEdu$Value) #format as numeric

# merge map and data
eurEduMapDf <- merge(eurMapDf, eurEdu, by.x="id", by.y="GEO")
eurEduMapDf <- eurEduMapDf[order(eurEduMapDf$order),]

eurEduMapDf <- subset(eurEduMapDf, 
                      long > min(europe.limits$lon) & long < max(europe.limits$lon) & 
                      lat  > min(europe.limits$lat) & lat  < max(europe.limits$lat))

# ggplot mapping
# data layer
m0 <- ggplot(data=eurEduMapDf)
# empty map (only borders)
m1 <- m0 + geom_path(aes(x=long, y=lat, group=group), color='gray') + coord_equal()

# fill with education expenditure data
m2 <- m1 + geom_polygon(aes(x=long, y=lat, group=group, fill=Value))

# inverse order (to have visible borders)
m0 <- ggplot(data=eurEduMapDf)
m1 <- m0 + geom_polygon(aes(x=long, y=lat, group=group, fill=Value)) + coord_equal()
m2 <- m1 + geom_path(aes(x=long, y=lat, group=group), color='black')
m2

# over a GoogleMap (not working if not correctly projected)
map <- get_map(location = 'Europe', zoom=4)
m0 <- ggmap(map)
m1 <- m0 + geom_polygon(aes(x=long, y=lat, group=group, fill=Value), data=eurEduMapDf, alpha=.9)
m2 <- m1 + geom_path(aes(x=long, y=lat, group=group), data=eurEduMapDf, color='black')

# add text
library(doBy) # from http://cran.r-project.org/web/packages/doBy/index.html
txtVal <- summaryBy(long + lat + Value ~ id, data=eurEduMapDf, FUN=mean, keep.names=T)
m3 <- m2 + geom_text(aes(x=long, y=lat, label=Value), data=txtVal, col="yellow", cex=3)
m3
```

## Maps and functions for Portugal

```{r}
PT_limits <- geocode(c(    # from http://pt.wikipedia.org/wiki/Pontos_extremos_de_Portugal
  "Cristoval,Portugal",                     # mais setentrional
  "Paradela,Portugal",                      # mais oriental
  "Cabo da Roca, Sintra,Portugal",          # mais ocidental
  "Cabo de Santa Maria,Algarve,Portugal")   # mais meridional
)

plot(eurMap, xlim = range(PT_limits$lon), ylim = range(PT_limits$lat), asp = 1.0)
```

```{r, echo=FALSE}
# ptMap <- readShapePoly(fn="Cont_AAD_CAOP2014")
# plot(ptMap)
# 
# ptMapDf <- fortify(ptMap)
# tail(ptMapDf)
# 
# 
# ptMapTroco <- readShapeLines(fn="Cont_Troco_CAOP2014")
# plot(ptMapTroco)
# ptMapTrocoDf <- fortify(ptMapTroco)
# head(ptMapTrocoDf)
```



```{r}
# pt_map <- get_map(location = 'Portugal', zoom = 7, language = "PT-pt")
# this is a better fit:
pt_map <- get_map(location = c(lon=mean(PT_limits$lon), lat= mean(PT_limits$lat)), zoom = 7, language = "PT-pt", maptype="satellite")

ggmap(pt_map)

# http://www.dgterritorio.pt/cartografia_e_geodesia/cartografia/carta_administrativa_oficial_de_portugal__caop_/caop_em_vigor/
  
  
```

```{r}
freguesias <- read.csv("freguesias.csv", header = TRUE)
head(freguesias)
municipios <- read.csv("municipios.csv", header = TRUE)
head(municipios)
```


##############################################

Functions for mapping dstrict and municipe information of Portugal

The maps from Portugal were from [Global Administrative Areas](http://gadm.org/country) datafiles. Zip file with shapefiles from [Portugal](http://biogeo.ucdavis.edu/data/gadm2/shp/PRT_adm.zip)

Ref: [http://statisticaconr.blogspot.it/2009/12/choropleth-map-in-r-coloriamo-le-mappe.html](http://statisticaconr.blogspot.it/2009/12/choropleth-map-in-r-coloriamo-le-mappe.html) (In Italian)

```{r, fig.height=12, fig.width=8}
library(maptools) # read shapefiles (.shp)

# district_data is a data frame with two columns, district and value
pt_distritos <- function(district_data, breaks, palette = "hot", 
                         show_label=TRUE, 
                         # the next args define how to handle non-available datapoints
                         show.NA=FALSE, val.NA=-1, col.NA="black", label.NA="NA") {
  
  # read shapefile file for districts
  library(maptools)
  ptMap <- readShapePoly(fn="PRT_adm1")
  
  init <- ifelse(show.NA,val.NA,min(municipe_data[,2]))

  df <- data.frame(districts = ptMap$NAME_1,
                   value     = rep(init,length(ptMap$NAME_1)))

  # populate df with district_data (the districts might be in different sorting order)
  for(i in 1:nrow(df)) {
    query <- as.character(district_data[,1])==as.character(df[i,1])
    if (any(query))
      df[i,2] = district_data[which(query),2]
  }
  
  if (any(df==-1))
    breaks = c(0,breaks)
  
  library(Hmisc)
  lev <- cut2(df$value, cuts=breaks)
  n.cols <- length(breaks)+1

  if(show_label && length(breaks)>3) { # does not work with small number of breaks (?)
    label_names <- levels(lev)
    if (show.NA && any(df==-1))
      label_names[1] <- "NA"
  } else
    label_names <- ""
  
  if (palette=="hot") {
    cols <- heat.colors(n.cols)
    if (show.NA && any(df==-1))
      cols <- c(col.NA,cols)
  }
  else if (palette == "grey")
    cols <- grey(seq(0, 1, length = n.cols))
  else if (palette == "rainbow") {
    cols <- rainbow(n.cols)
    if (show.NA && any(df==-1))
      cols <- c(col.NA,cols)
  }
 
  ptMap$value <- as.factor( as.character(as.numeric(lev)) )
  
  library(sp)
  spplot(ptMap, "value", col.regions=cols, 
         xlim = range(-10,-6), ylim = range(36.9,42.2), asp = 1.0,
         colorkey = list(labels = list(labels = label_names, width = 1, cex = 1),
                         space = "bottom"))
}

# Use eg:
library(maptools)
ptMap <- readShapePoly(fn="PRT_adm1")
district_data <- data.frame(district = ptMap$NAME_1, 
                            value    = rpois(20,10)) # make some fake data
district_data <- district_data[1:19,] # remove one district, for testing

pt_distritos(district_data, breaks=c(3,6,11,15), palette="rainbow", show.NA=TRUE)
```

And the same for the municipal areas:

```{r, fig.height=12, fig.width=8}
# district_data is a data frame with two columns, municipe and value
pt_municipios <- function(municipe_data, breaks, palette = "hot", 
                          show_label=TRUE, 
                          # the next args define how to handle non-available datapoints
                          show.NA=FALSE, val.NA=-1, col.NA="black", label.NA="NA") {
  
  # read shapefile file for districts
  library(maptools)
  ptMap <- readShapePoly(fn="PRT_adm2")
  
  init <- ifelse(show.NA,val.NA,min(municipe_data[,2]))
  
  df <- data.frame(municipios = ptMap$NAME_2,
                   value      = rep(init,length(ptMap$NAME_2)))

  # populate df with municipe_Data (the municipes might be in different sorting order)
  for(i in 1:nrow(df)) {
    query <- as.character(municipe_data[,1])==as.character(df[i,1])
    if (any(query))
      df[i,2] = municipe_data[which(query),2]
  }
  
  if (any(df==-1))
    breaks = c(0,breaks)
  
  library(Hmisc)
  lev <- cut2(df$value, cuts=breaks)
  n.cols <- length(breaks)+1

  if(show_label && length(breaks)>3) { # does not work with small number of breaks (?)
    label_names <- levels(lev)
    if (show.NA && any(df==-1))
      label_names[1] <- "NA"
  } else
    label_names <- ""
  
  if (palette=="hot") {
    cols <- heat.colors(n.cols)
    if (show.NA && any(df==-1))
      cols <- c(col.NA,cols)
  }
  else if (palette == "grey")
    cols <- grey(seq(0, 1, length = n.cols))
  else if (palette == "rainbow") {
    cols <- rainbow(n.cols)
    if (show.NA && any(df==-1))
      cols <- c(col.NA,cols)
  }
  
  ptMap$value <- as.factor( as.character(as.numeric(lev)) )
  
  library(sp)
  spplot(ptMap, "value", col.regions=cols, 
         xlim = range(-10,-6), ylim = range(36.9,42.2), asp = 1.0,
         colorkey = list(labels = list(labels = label_names, width = 1, cex = 1),
                         space = "bottom"))
}

# Use eg:
municipe_data <- read.csv("passivos.csv", header = TRUE)
head(municipe_data)

pt_municipios(municipe_data, breaks=c(1e2,1e3,5e3,1e4,2e4), palette="hot", show.NA=TRUE)
```

Also check [https://procomun.wordpress.com/2012/02/18/maps_with_r_1/](https://procomun.wordpress.com/2012/02/18/maps_with_r_1/)
and [https://procomun.wordpress.com/2012/02/20/maps_with_r_2/](https://procomun.wordpress.com/2012/02/20/maps_with_r_2/).