---
title: "Empirical Entropy"
author: "João Neto"
date: "October 2020"
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 12
    fig_height: 6
---

```{r}
library(magrittr)
```


Probability distributions are functions that assign probability mass or density to a universe of outcomes, which are useful to model random events. Different distributions have different mappings from events to probabilities, and some are more uncertain than others.

Since entropy is a measure of uncertainty, it is to no surprise that 
distributions have an associated entropy.

## Entropy for Discrete Distributions

For discrete distributions, over all values $x$ inside the domain, the entropy $H$ is given by expression

$$H = -\sum_x p(x) \log p(x)$$

Consider the example of the Poisson where

$$p(k~|~\lambda) = \frac{\lambda^k e^{-\lambda}}{k!}$$

the analytic solution is

$$\lambda (1- \log \lambda) + e^{-\lambda}\sum_{k=0}^\infty \frac{\lambda^k \log k!}{k!}$$

The next function approximates this expression

```{r}
entropy.poisson <- function(lambda) {
  lambda*(1-log(lambda)) + 
    exp(-lambda) * 
    sum(sapply(0:20, 
               function(k) lambda^k * log(factorial(k)) / factorial(k)))
}
```

Say, for $\lambda=5$ the entropy will be

```{r}
lambda <- 5
entropy.poisson(lambda)
```

## Empirical Entropy for Discrete Distributions

What if we have a discrete sample taken from some distribution?

```{r}
lambda <- 5
xs1 <- rpois(1e6, lambda)
head(xs1, 10)
```

If we know which distribution the sample was taken, we can infer the parameters and then apply the respective entropy formula.

In our case, `xs1` was the result of a Poisson sampling, so:

```{r, message=FALSE}
lambda.hat <- mean(xs1)
entropy.poisson(lambda.hat)
```

But if we don't have a clue about the underlining distribution?

The first step is to compute the empirical point probabilities

$$\hat{p}(x_i) = \frac{1}{n} \sum_{k=1}^n \delta_{x_i}(x_k)$$

where $\delta_{x_i}(x_k)$ is 1 if $x_i = x_k$, zero otherwise.

The empirical entropy is then

$$\hat{H} = -\sum_x \hat{p}(x) \log \hat{p}(x)$$

The next R function computes this value,

```{r}
# ref: https://stats.stackexchange.com/questions/28178
empirical.entropy <- function(xs) {
  # empirical point probabilities
  epps <- (table(xs) / length(xs)) %>%  as.vector()
  
  -sum(epps * log(epps))
}
```

Let's check the true entropy with the empirical entropy of the Poisson sample:

```{r, collapse=TRUE}
entropy.poisson(lambda) # true entropy
empirical.entropy(xs1)  # empirical entropy
```

Another example, this time with a discrete uniform $\mathcal{U}\{a,b\}$, with theoretical entropy $\log (b-a+1)$,


```{r, collapse=TRUE}
xs2 <- sample(1:6, 1e6, rep=TRUE)  # uniform U{1,6}

log(6-1+1)             # true entropy
empirical.entropy(xs2) # empirical entropy
```

## Entropy for Continuous Distributions

Claude Shannon tried to generalize his entropy formula for the continuous case simply replacing the sum with an integral,

$$H = - \int_x f(x) \log(x) dx$$

but without any mathematical derivation. This is denoted as _differential entropy_.

This formula lacks many properties of the discrete case.

For the classic continuous distributions, their entropy is known. For the Gaussian $\mathcal{N}(\mu, \sigma$ the entropy is $\frac{1}{2}\log(2\pi e \sigma^2)$

```{r}
entropy.normal <- function(mu, sigma) {
  0.5 * log(2*pi*exp(1)*sigma^2)
}
```

Let's check an example,

```{r}
mu <- 0
sigma <- 0.35

entropy.normal(mu, sigma)
```

ET Jaynes [argued](https://en.wikipedia.org/wiki/Limiting_density_of_discrete_points) that the continuous case should be defined as the limiting case of increasingly dense discrete distributions.

## Empirical Entropy for Continuous Distributions

One approximation is given by producing an histogram of the continuous values, and then accounting for the values $x_i$ inside bin $i$ with  width $w_i$,

$$\hat{H} = -\sum_i \hat{p}(x_i) \log \frac{\hat{p}(x_i)}{w_i}$$

In R:

```{r}
# ref: https://en.wikipedia.org/wiki/Entropy_estimation
empirical.entropy.continuous <- function(xs, breaks=1e3) {
  range.xs   <- max(xs)-min(xs)
  width.bins <- range.xs/breaks # all bins have the same width

  # bin the results into finite intervals
  xs <- xs %>% cut(breaks=breaks) %>% as.numeric()
  # empirical point probabilities
  epps <- (table(xs) / length(xs)) %>%  as.vector()
  
  -sum(epps * (log(epps) - log(width.bins)))
}
```

Let's compare the true and empirical entropy for a normal sample,

```{r, collapse=TRUE}
n <- 1e6
xs3 <- rnorm(n, 0, sigma)

entropy.normal(mu, sigma)         # true entropy
empirical.entropy.continuous(xs3) # empirical entropy
```


