Binomial Likelihood with Hierarchical Prior
========================================================

This file codes and plots figure 9.2 (p.162) from 
Kruschke - Doing Bayesian Data Analysis (2010)

```{r init, results='hide', message=FALSE}
library(aws)
library(lattice) 
library(base)

rotate       <- (-25)  # parameters used for 3D plots
tilt         <- 25
parallelness <- 5.0
shadeval     <- 0.05
perspcex     <- 0.7
ncontours    <- 9
```

First we plot the Prior distribution $p(\theta,\mu)$. This prior is calculated
by the product of the likelihood of the parameter $\theta$ given hyperparameter $\mu$, i.e., $p(\theta|\mu)$ by the prior $p(\mu)$. So,  $p(\theta,\mu) = p(\theta|\mu) \times p(\mu)$. The prior $p(\mu)$ is modeled by a $beta(2,2)$ and the likelihood by a $beta(100\mu, 100(1-\mu))$. The 100 in the last formula means we assign high confidence to the dependence of $\theta$ on the hyperparameter $\mu$.

```{r prior, fig.width=12, fig.height=6, dependson="init"}
# make a grid of thetas and mus
dim.grid <- 100  # arbitrary number of intervals for grid on theta.
thetas   <- seq(from=((1/dim.grid)/2), to=(1-((1/dim.grid)/2)), by=(1/dim.grid))
mus      <- thetas

# the prior function, muiltiplying the two beta distributions
fPrior <- function(theta,mu) {
  return(dbeta(theta,100*mu,100*(1-mu)) * dbeta(mu,2,2))
}

priors <- outer(thetas, mus, fPrior)  # outer product
priors <- priors / sum(priors)        # normalize

length.step <- length(thetas)
num.steps   <- 30
step.thetas <- seq( 1, length.step, by = round( length.step/num.steps ) )
step.mus    <- step.thetas

zmax <- max(priors)
par(mfrow=c(1,2))
persp(thetas[step.thetas] , mus[step.mus] , priors[step.thetas,step.mus],
      xlab="theta" , ylab="mu" , main="Prior Distribution" , cex=perspcex ,
      lwd=0.1  , xlim=c(0,1) , ylim=c(0,1) , zlim=c(0,2*zmax),
      zlab="p(theta,mu)" , theta=rotate , phi=tilt , d=parallelness ,
      shade=shadeval)

# make the contour plot
contour(thetas[step.thetas] , mus[step.mus] , priors[step.thetas,step.mus],
        main=bquote(" ") , levels=signif(seq(0,zmax,length=ncontours),3) ,
        drawlabels=FALSE , xlab=bquote(theta) , ylab=bquote(mu) )

par(mfrow=c(1,1))

# check a slice of the prior distribution
mu <- 0.25
col <- 100*round(mu*100/dim.grid,2) # find right column

plot(priors[,col]/sum(priors[,col]),
     type="l",xlab=bquote(theta),
     ylab=bquote(paste("p(",theta,"|",mu,"=",.(mu),")")), 
     main=bquote(paste("A slice of p(",theta,"|",mu,")")))

```
The marginal probability $p(\mu) \approx \sum_{i} p(\theta_i, \mu)$

```{r marginal, dependson="prior"}
# compute marginal probability of p(mu) based on the prior grid
marginal.mu <- rep(0,dim.grid)
for(i in 1:dim.grid) {
  marginal.mu[i] <- sum(priors[,i]) 
}

# normalize
sum.priors <- sum(marginal.mu)
for(i in 1:dim.grid) {
  marginal.mu[i] <- marginal.mu[i] / sum.priors 
}
                                              
plot(marginal.mu, xlab=bquote(mu), ylab=bquote(p(mu)),type="l")
title("Marginal Probability p(mu)")
abline(v=seq(0,100,20),h=seq(.0025,.015,.0025),col="lightgray")

#compute marginal probability of p(theta) based on the prior grid
marginal.theta <- rep(0,dim.grid)
for(i in 1:dim.grid) {
  marginal.theta[i] <- sum(priors[i,]) 
}

# normalize
sum.priors <- sum(marginal.theta)
for(i in 1:dim.grid) {
  marginal.theta[i] <- marginal.theta[i] / sum.priors 
}
                                              
plot(marginal.theta, xlab=bquote(theta), ylab=bquote(p(theta)),type="l")
title("Marginal Probability p(theta)")
abline(v=seq(0,100,20),h=seq(.002,.016,.002),col="lightgray")
```

Now comes some data D. We had 12 coin drops and 9 were heads.

This means the likelihood function $p(D|\mu,\theta) = p(D|\theta) = \theta^9 \times (1-\theta)^3$ ($\theta$ is the probability for heads).

Let's plot it:

```{r likelihood, dependson="prior"}
n.heads <- 9
n.tails <- 3

# currying function
Curry <- function(FUN,...) { 
  .orig <- list(...)
  function(...) {
    do.call(FUN, c(.orig, list(...)))
  }
}

likelihood <- function(theta, mu, heads, tails) {
  return (theta^heads * (1-theta)^tails)
}

data.likelihood <- Curry(likelihood, heads=n.heads, tails=n.tails)

likelihoods <- outer(thetas, mus, data.likelihood)  # outer product

length.step <- length(thetas)
num.steps   <- 30
step.thetas <- seq( 1, length.step, by = round( length.step/num.steps ) )
step.mus    <- step.thetas

zmax <- max(likelihoods)
persp(thetas[step.thetas] , mus[step.mus] , likelihoods[step.thetas,step.mus],
      xlab="theta" , ylab="mu" , main="Likelihood Distribution", cex=perspcex,
      lwd=0.1  , xlim=c(0,1) , ylim=c(0,1) , zlim=c(0,2*zmax),
      zlab="p(D|theta,mu)" , theta=rotate , phi=tilt , d=parallelness ,
      shade=shadeval)

```

Given the prior and the likelihood, we have what it takes to compute the posterior $p(\mu,\theta|D)$. The posterior is proportional to the prior times the likelihood, $p(\mu,\theta|D) \propto p(D|\theta,\mu) \times p(\theta,\mu) = p(D|\theta) \times p(\theta | \mu) \times p(\mu)$

```{r posterior, dependson=c("prior","likelihood")}
posteriors <- priors * likelihoods # simply as that!
evidence <- sum(posteriors)
posteriors  <- posteriors / evidence #normalize

length.step <- length(thetas)
num.steps   <- 30
step.thetas <- seq( 1, length.step, by = round( length.step/num.steps ) )
step.mus    <- step.thetas

zmax <- max(posteriors)
persp(thetas[step.thetas] , mus[step.mus] , posteriors[step.thetas,step.mus],
      xlab="theta" , ylab="mu" , main="Posterior Distribution", 
      sub=paste0("Evidence, p(D) = ", signif(evidence,3)),
      cex=perspcex, lwd=0.1  , xlim=c(0,1) , ylim=c(0,1) , zlim=c(0,2*zmax),
      zlab="p(theta,mu|D)" , theta=rotate , phi=tilt , d=parallelness ,
      shade=shadeval)

#compute marginal probability of p(mu|D) based on the posterior grid
marginal.post.mu <- rep(0,dim.grid)
for(i in 1:dim.grid) {
  marginal.post.mu[i] <- sum(posteriors[,i]) 
}

# normalize
sum.priors <- sum(marginal.post.mu)
for(i in 1:dim.grid) {
  marginal.post.mu[i] <- marginal.post.mu[i] / sum.priors 
}
                                              
plot(marginal.post.mu, xlab=bquote(mu), ylab=bquote(paste("p(",mu,"|D)")),
     type="l")
#title(expression("Marginal Distribution p("~theta~"|D)")) # cf demo(plotmath)
title(bquote(paste("Marginal Distribution p(",mu,"|D)")))
abline(v=seq(0,100,20),h=seq(.005,.035,.005),col="lightgray")

#compute marginal probability of p(theta|D) based on the posterior grid
marginal.post.theta <- rep(0,dim.grid)
for(i in 1:dim.grid) {
  marginal.post.theta[i] <- sum(posteriors[i,]) 
}

# normalize
sum.priors <- sum(marginal.post.theta)
for(i in 1:dim.grid) {
  marginal.post.theta[i] <- marginal.post.theta[i] / sum.priors 
}
                                              
plot(marginal.post.theta, xlab=bquote(theta), 
     ylab=bquote(paste("p(",theta,"|D)")),type="l")
#title(expression("Marginal Distribution p("~theta~"|D)")) # cf demo(plotmath)
title(bquote(paste("Marginal Distribution p(",theta,"|D)")))
abline(v=seq(0,100,20),h=seq(.005,.035,.005),col="lightgray")

```