---
title: "Color Palette"
author: "João Neto"
date: "January 2015"
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 4
    fig_height: 4
cache: TRUE
---

# R base

R provides with some base palettes:

+ rainbow

+ heat.colors

+ terreain.colors

+ topo.colors

+ cm.colors

```{r}
show_palette <- function(colors) {
  image(1:n, 1, as.matrix(1:n), col = colors, 
    xlab = "", ylab = "", xaxt = "n", 
    yaxt = "n", bty = "n")
}

n <- 6
show_palette(rainbow(n))
show_palette(heat.colors(n))
show_palette(terrain.colors(n))
show_palette(topo.colors(n))
show_palette(cm.colors(n))
```

There also a `alpha` parameter, from `[0,1]` for transparency.

```{r, fig.width=8}
par(mfrow=c(1,2))
show_palette(rainbow(n, alpha=0.33))
show_palette(rainbow(n, alpha =0.66))
```

For greyscales:

```{r}
n <- 20
greys <- grey(seq(0, 1, length = n))
show_palette(greys)
```


# RColorBrewer

[Ref](htpp://mran.revolutionanalytics.com/packages/info/?RColorBrewer)

> Cindy Brewer [website](http://colorbrewer2.org/) helps you choose the appropriate color scale for you map depending on your data type: qualitative, sequential or diverging (with a neutral color between two extremes).
 
There are 3 types of palettes, sequential, diverging, and qualitative.

+ Sequential palettes are suited to ordered data that progress from low to high

+ Diverging palettes put equal emphasis on mid-range critical values and extremes at both ends of the data range

+ Qualitative palettes are best suited to representing nominal or categorical data

```{r, fig.width=8, fig.height=8}
# install: install.packages('RColorBrewer')
library(RColorBrewer)

display.brewer.all() 
```

To get the colors of a given palette:

```{r}
brewer.pal(11,"Spectral")
display.brewer.pal(11,"Spectral")
```

ggplot2 can use these palettes via the [scale_brewer](http://docs.ggplot2.org/current/scale_brewer.html) option.

