---
title: "DiagrammeR examples"
date: "May, 2015"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

Refs: 

+ [http://rich-iannone.github.io/DiagrammeR/graphs.html](http://rich-iannone.github.io/DiagrammeR/graphs.html)

+ [http://www.graphviz.org/pdf/dotguide.pdf](http://www.graphviz.org/pdf/dotguide.pdf)

```{r}
library(DiagrammeR)
```

Some complex eg to show several types of tools:

```{r}
grViz("
  digraph boxes_and_circles {
  
    # a 'graph' statement
    graph [overlap = true, fontsize = 10]
  
    # several 'node' statements
    node [shape = box,
          fontname = Helvetica]
    A; B; C; D; E; F
  
    node [shape = circle,
          fixedsize = true,
          width = 0.9] // sets as circles
    2; 3; 4; 5; 6; 7

    node [shape = oval,
          fixedsize = true,
          peripheries = 2,
          width = 0.9,
          fontname = 'Courier New',
          style=filled,
          fillcolor = red]
    8

    node [label =<<table border='0' cellborder='1' cellpadding='3' bgcolor='white'>
                    <tr><td bgcolor='black' align='center' colspan='2'><font color='white'>header</font> </td></tr>
                    <tr><td align='left' port='r2'>&#40;2&#41; &#188; &bull;</td>
                        <td bgcolor='grey' align='right'>$</td>
                    </tr>
                  </table>>
          ] // sets as circles
    1

    # several 'edge' statements
    A->1 B->2 B->3 B->4 C->A
    1->D E->A 2->4 1->5 1->F
    E->6 [label='x &#x2208; 3']
    4->6 [label='&#946;&#x2082;=0.85']
    5->7 [penwidth = 5]
    6->7 [style=dotted,label='dotted']
    3->8 [headport = 'e', tailport = 'e', color=blue]
  }
")
```

The greek letters (Cf. [here](http://www.wizcity.com/Computers/Characters/GreekUTF8.php):

`&Alpha;  &Beta;  &Gamma;  &Delta;  &Epsilon; &Zeta;  &Eta;  &Theta;  &Iota;  &Kappa;`  
`&Lambda; &Mu;  &Nu;  &Xi;  &Omicron; &Pi;  &Rho;  &Sigma;  &Tau;  &Upsilon; &Phi; `
`&Chi;  &Psi;  &Omega;  &alpha;  &beta;  &gamma;  &delta;  &epsilon; &zeta;  &eta; `
`&theta;  &iota;  &kappa;  &lambda; &mu;  &nu;  &xi;  &omicron; &pi;  &rho;  &sigmaf;`
`&sigma;  &tau;  &upsilon; &phi;  &chi;  &psi;  &omega;  &thetasym &upsih;  &piv;`

Superscripts and Subscripts are the block U+2070 to U+209F

More symbols: [http://csbruce.com/software/utf-8.html](http://csbruce.com/software/utf-8.html). To translate, eg, `u+2208` is `&#x2208;`.

This is a eg for a probabilistic graphical model description:

```{r}
grViz("
  digraph dot {

    graph [compound = true, nodesep = .5, ranksep = .25,
           color = crimson, label='Polynomial Regression Model'
           /*, rankdir='LR', style=filled, fillcolor = blue*/
          ]

    subgraph cluster1 {
      node [shape = diamond, 
            color = black]
      xi

      node [shape = circle,
            style = filled,
            fillcolor = grey] 
      ti
      
      edge [color = black] 
      xi -> ti

      label='N'
    }  
  
    node [shape = circle]
    w

    node [shape = diamond,
          color = black,
          label = '&alpha;']
    alpha

    node [shape = diamond,
          color = black,
          label = '&sigma;&#x00B2;']
    sigma

    edge [color = black]
    alpha -> w
    w -> ti
    sigma -> ti 

  }
  ",
engine = "dot")
```

where diamonds describe deterministic parameters; shaded circles describe observed values; and boxes describe multiple variables (in this case $x_1 \ldots x_n$ and $t_1 \ldots t_n$).


