---
title: "Image Manipulation with EBImage"
author: "João Neto"
date: "February 2015"
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 4
---

Ref:

+ [Introduction to EBImage](http://www.bioconductor.org/packages/release/bioc/vignettes/EBImage/inst/doc/EBImage-introduction.pdf)

+ [Image Analysis using EBImage](http://www.r-bloggers.com/r-image-analysis-using-ebimage/)

From the authors:

> EBImage is an R package which provides general purpose functionality for the reading, writing, processing and analysis of images. 

```{r}
# to install: source("http://bioconductor.org/biocLite.R"); biocLite("EBImage")
library("EBImage")

pic <- readImage("sicily.jpeg")
display(pic)
```

Image Properties

```{r}
dim(pic)    # there are 3 matrices for RGB colors
print(pic)  # the RGB values are from [0,1]
```

Change to Greyscale

```{r}
pic_bw <- pic
colorMode(pic_bw) <- Grayscale
display(pic_bw)
```

Adjust Brigthness

```{r}
pic1 <- pic + 0.1 # more light
pic2 <- pic - 0.1 # less light
par(mfrow=c(1,2))
display(pic1)
display(pic2)
```

Adjust Contrast

```{r}
pic1 <- pic * 0.5 # reduce contrast
pic2 <- pic * 2   # increase contrast
par(mfrow=c(1,2))
display(pic1)
display(pic2)
```

Gamma Correction

```{r}
pic1 <- pic ^ 2 
pic2 <- pic ^0.7 
par(mfrow=c(1,2))
display(pic1)
display(pic2)
```

Cropping

```{r}
display(pic[100:200,100:300,])
```

Rotating and Translating

```{r}
pic1 <- rotate(pic, 45)
pic2 <- translate(rotate(pic, 45), c(150, 50))
par(mfrow=c(1,2))
display(pic1)
display(pic2)
```

Other operations

```{r}
par(mfrow=c(3,2))
display(pic > 0.5)
display(transpose(pic))
display(flip(pic))
display(flop(pic))
display(resize(pic,300,300))
# applying a matrix affine transformation (check help file)
m <- matrix(c(0.6, 0.2, 0, -0.2, 0.3, 300), nrow=3)
display( affine(pic, m) )
```

Changing color layers

```{r}
imgk    <- channel(pic, 'rgb')
imgk[236:276, 106:146, 1] = 1
imgk[236:276, 156:196, 2] = 1
imgk[236:276, 206:246, 3] = 1
display(imgk)

# make a color square inside a BW picture
imgk    <- channel(pic, 'rgb')
imgk_bw <- channel(pic_bw, 'rgb')
r <- imgk_bw[,,,1]
g <- imgk_bw[,,,1]
b <- imgk_bw[,,,1]
r[100:200, 200:300,] <- imgk[100:200, 200:300, 1]
g[100:200, 200:300,] <- imgk[100:200, 200:300, 2]
b[100:200, 200:300,] <- imgk[100:200, 200:300, 3]
display(rgbImage(red=r, green=g, blue=b)) 
```

Filtering

```{r}
filter_low <- makeBrush(21, shape= 'disc', step=FALSE)^2
filter_low <- filter_low/sum(filter_low)
pic_low <- filter2(pic, filter_low)  # low pass, ie, blur
display(pic_low)

filter_high <- matrix(1, nc = 3, nr = 3)
filter_high[2, 2] <- -8
pic_high <- filter2(pic, filter_high)
display(pic_high)   # high pass

pic_med <- medianFilter(pic, 1.1)  # median filter  
display(pic_med)   
```

