---
title: "Using ggplot2"
date: "July 2015"
output: 
  html_document:
    fig_width: 12
    toc: true
    toc_depth: 3
---

Refs:

+ Hadley Wichham - *ggplot2, Elegant Graphics for Data Analysis*

+ Winston Chang - [R Graphics Cookbook](http://www.cookbook-r.com/Graphs/index.html)

+ [ggplot2 online documentation](http://docs.ggplot2.org/current/)

ggplot2 implements Wilkinson grammar of graphics that describes and conceptually organizes the features that underlie statistical graphics.

The most important concepts are:

+ The **data** (duh!)

+ **Aesthetic** mappings describing how the data is mapped to aesthetic atributes (eg, shape, color, horizontal and vertical position)

+ Geometric objects, **geoms**, that we see on the graphic (eg, points, lines)

+ Statistical mappings, **stats**, summarizing the data (eg, binning, counting)

+ Scaling mappings, **scales**, the map the data space into aesthetic space (eg, linear or log axis, legends, how to map colors), there should be a scale for each aesthetic

+ Coordinate systems, **coord**, a map from the data coordinates to the graph coordinates (eg, cartesian, log-log, polar)

+ **Facets** describing how to plit the data into subsets

```{r, comment=FALSE}
library(ggplot2)
library(gridExtra)  # for presenting plots side by side

set.seed(101)
d <- diamonds[sample(nrow(diamonds), 100), ]  # read some data for next examples
head(d)
```


## `qplot`

`qplot` is ggplot2's plot function.

```{r}
plot1 <- qplot(carat,      price,      data = d)
plot2 <- qplot(log(carat), log(price), data = d)
plot3 <- qplot(carat,      x * y * z,  data = d)  # x*y*z gives the volume
grid.arrange(plot1, plot2, plot3, ncol=3)
```

`qplot` automates some aesthetics like how to assign colors and shapes to data:

```{r}
plot1 <- qplot(carat, price, data = d, colour = color, size= carat)
plot2 <- qplot(carat, price, data = d, shape  = cut)
grid.arrange(plot1, plot2, ncol=2)
```

Notice that the plots also come with a legend. For each aesthetic attribute there is a scale function mapping the data values to the aesthetic values. Eg, in the left plot, the attribute D was associated with red.

`qplot` accepts different types of geometric objects, geoms, which will make it produce different types of graphics. The default is `geom="point"`, ie, the scatterplots we've seen. Other objects are possible:

```{r}
plot1 <- qplot(carat, price, data = d, geom=c("point", "smooth")) # default smooth by loess regression
plot2 <- qplot(carat, price, data = d, geom=c("point", "boxplot"))
plot3 <- qplot(carat, price, data = d, geom=c("point", "line")) 
grid.arrange(plot1, plot2, plot3, ncol=3)
```

Geoms for 1D data:

```{r}
plot1 <- qplot(carat, data = d, geom="histogram") # continuous values
plot2 <- qplot(color, data = d, geom="bar")       # discrete values
grid.arrange(plot1, plot2, ncol=2)  
plot1 <- qplot(carat, data = d, geom="freqpoly")
plot2 <- qplot(carat, data = d, geom="density") 
grid.arrange(plot1, plot2, ncol=2)  
```

### Smooth geoms

The smooth geom can be used with different regression methods:

```{r}
plot1 <- qplot(carat, price, data = d, geom=c("point", "smooth"), method="lm")
plot2 <- qplot(carat, price, data = d, geom=c("point", "smooth"), method="lm", formula=y~poly(x,3)) # polynomial regression
grid.arrange(plot1, plot2, ncol=2)
library(splines) # using natural splines
plot3 <- qplot(carat, price, data = d, geom=c("point", "smooth"), method="lm", formula=y~ns(x,5)) 
library(MASS) # for robust regression
plot4 <- qplot(carat, price, data = d, geom=c("point", "smooth"), method="rlm") 
grid.arrange(plot3, plot4, ncol=2)
```

### Jitter and opacity

Sometimes the datapoints are too many and a direct plot is unable to transmit an appropriate perspective of the data. One tool is to jitter the points (add small random noise so that many equal data points are spread around its center) and/or define an amount of opacity, ie, stating how many points there must be at area so that the graphic plots without transparency.

```{r}
plot1 <- qplot(carat, price, data = diamonds)
plot2 <- qplot(carat, price, data = diamonds, alpha=I(1/50)) # 100 pts for total opacity
grid.arrange(plot1, plot2, ncol=2)

plot1 <- qplot(color, price/carat, data = diamonds)
plot2 <- qplot(color, price/carat, data = diamonds, geom = "jitter")
plot3 <- qplot(color, price/carat, data = diamonds, geom = "jitter", alpha = I(1/10))
grid.arrange(plot1, plot2, plot3, ncol=3)
```

### Histograms and density plots

For the histogram geom we can define the bin size:

```{r}
plot1 <- qplot(carat, data = diamonds, geom = "histogram", binwidth = 1)
plot2 <- qplot(carat, data = diamonds, geom = "histogram", binwidth = 0.1)
plot3 <- qplot(carat, data = diamonds, geom = "histogram", binwidth = 0.05)
grid.arrange(plot1, plot2, plot3, ncol=3)
```

For density plots we have the adjust parameter:

```{r}
plot1 <- qplot(carat, data = diamonds, geom = "density", adjust = 4)
plot2 <- qplot(carat, data = diamonds, geom = "density", adjust = 1)
plot3 <- qplot(carat, data = diamonds, geom = "density", adjust = 0.5)
grid.arrange(plot1, plot2, plot3, ncol=3)
```

Applying a color aesthetic:

```{r}
plot1 <- qplot(carat, data = diamonds, geom = "histogram", fill = color)
plot2 <- qplot(carat, data = diamonds, geom = "density", colour = color)
grid.arrange(plot1, plot2, ncol=2)
```

### Time series

```{r, comment=FALSE}
head(economics)
year <- function(x) as.POSIXlt(x)$year + 1900
economics$year <- year(economics$date)
head(economics)
```

We can use geom line for standard time series:

```{r}
plot1 <- qplot(date, unemploy / pop, data = economics, geom = "line")
plot2 <- qplot(date, uempmed,        data = economics, geom = "line")
grid.arrange(plot1, plot2, ncol=2)
```

The geom path that joins points adjacent in time (when scatterplots does not provide enough information):

```{r}
plot1 <- qplot(unemploy/pop, uempmed, data = economics, geom = c("point", "path"))
plot2 <- qplot(unemploy/pop, uempmed, data = economics, geom = c("point", "path"), color=year)
grid.arrange(plot1, plot2, ncol=2)
```

### Faceting

Faceting splits the data into subsets which are present at different graphs for easier comparisation.

```{r, warning=FALSE}
plot1 <- qplot(carat, data = diamonds, facets = color ~ ., geom = "histogram", binwidth = 0.1,  xlim = c(0, 3))
plot2 <- qplot(carat, data = diamonds, facets = cut ~ .,   geom = "density",   binwidth = 0.05, xlim = c(0, 3))
grid.arrange(plot1, plot2, ncol=2)
```

```{r, warning=FALSE}
plot1 <- qplot(displ, hwy, data = mpg, facets = . ~ drv)
plot2 <- qplot(hwy,        data = mpg, facets = drv ~ ., binwidth = 2)
grid.arrange(plot1, plot2, ncol=2)
```


### Other options

Here's the use of some other parameters for `qplot`:

```{r, warning=FALSE}
qplot(carat, price, data=d, 
      xlim=c(0.5,1.5), ylim=c(0,5e3),
      main="Main Title",
      xlab=expression(beta[1] == 1), ylab="some stuff")
```

## `ggplot` and Layers

Layers are responsible for creating the objects that we perceive on the plot. A layer is composed of four parts:

+ data and aesthetic mapping,

+ a statistical transformation (stat),

+ a geometric object (geom)

+ a position adjustment.

`qplot` does everything inside its parameters. If we like to have more control in the creation of the graphic, we need to use `ggplot`.

`ggplot` receives two arguments, the data which must be a data frame and which aesthetics mappings we want. An eg:

```{r, error=TRUE}
p <- ggplot(d, aes(carat, price, colour=cut))
p
```

It gave an error because we still not defined geoms so that there is something to see.
To add new layers we should use operator `+`:

```{r}
plot1 <- p + layer(geom="point")
plot2 <- p + layer(geom="line")
grid.arrange(plot1, plot2, ncol=2)  
```

`ggplot2` has a series of specialized functions that simplify the use of `layer` with prefixes `geom_XXX` or `stat_XXX`

```{r}
p <- ggplot(diamonds, aes(x=carat)) 
plot1 <- p + layer(geom="bar", 
                   geom_params = list(fill = "steelblue"), 
                   stat = "bin", 
                   stat_params = list(binwidth = 0.25))
plot2 <- p + geom_histogram(binwidth = 0.25, fill = "steelblue")  # same graph
grid.arrange(plot1, plot2, ncol=2)  
```

We can add multiple layers:

```{r, warning=FALSE}
ggplot(msleep, aes(sleep_rem / sleep_total, awake)) +
  geom_point() + 
  geom_smooth()
```

### Data

iIt's possible to use the same graph objet with another data using operator `%+%`:

```{r}
head(mtcars)
plot1 <- ggplot(mtcars, aes(mpg, wt, colour = cyl)) + geom_point()
mtcars2 <- transform(mtcars, wt = wt ^ 2)
plot2 <- plot1 %+% mtcars2 
grid.arrange(plot1, plot2, ncol=2)  
```

### Aesthetics

Function `aes` describes which data values are assigned to things we see in the graph.
The first two parameters are to define which should be considered as `x` and `y`. Other mappings can be stated, like colour like in the previous eg `aes(mpg, wt, colour = cyl)`.

```{r}
p <- ggplot(mtcars) 
plot1 <- p + aes(wt,   hp)  + geom_point(colour="darkblue")             # set colors
plot2 <- p + aes(disp, mpg) + geom_point() + aes(colour = factor(cyl))  # map colors
grid.arrange(plot1, plot2, ncol=2) 
```

### Grouping

We can split the data into distinct groups, ie, datapoints that share something in common, and then plot them differently. This is done with the `group` mapping.

```{r}
plot1 <- ggplot(mtcars, aes(wt, hp, colour=factor(cyl), group=cyl)) + 
           geom_smooth(method="lm")
plot2 <- ggplot(mtcars, aes(wt, hp, colour=factor(cyl), group=cyl)) + 
           geom_boxplot() + theme(legend.position="none") # removes legend
grid.arrange(plot1, plot2, ncol=2)
```

### Geoms

These dtermine the rendering of the graph. Each geom has a set of aesthetics it understands and a default statistic.

Some egs:

```{r}
p <- ggplot(mtcars) + aes(wt, hp) + geom_point()
plot1 <- p + 
         geom_abline(intercept=2, slope=20, colour="red", size=1.5, linetype="dashed") +
         geom_hline(yintercept=seq(100, 150, by=15)) +
         geom_text(label="    A", angle=30)
plot2 <- p + geom_path(size = 0.5, lineend = "round")
plot3 <- p + geom_rect(xmin=2, xmax=3, ymin=200, ymax=300, fill="lightblue")
grid.arrange(plot1, plot2, plot3, ncol=3)
```

```{r}
df <- data.frame(x = c(3, 1, 5), y = c(2, 4, 6), label = c("a","b","c"))
p <- ggplot(df, aes(x, y, label = label)) + xlab(NULL) + ylab(NULL)
plot1 <- p + geom_point()   + ggtitle("geom_point")
plot2 <- p + geom_bar(stat="identity") + ggtitle("geom_bar(stat=\"identity\")")
plot3 <- p + geom_line()    + ggtitle("geom_line")
plot4 <- p + geom_area()    + ggtitle("geom_area")
plot5 <- p + geom_path()    + ggtitle("geom_path")
plot6 <- p + geom_text()    + ggtitle("geom_text")
plot7 <- p + geom_tile()    + ggtitle("geom_tile")
plot8 <- p + geom_polygon() + ggtitle("geom_polygon")
grid.arrange(plot1, plot2, plot3, plot4, plot5, plot6, plot7, plot8, nrow=2, ncol=4)
```

### Positions

Position adjustments modify the position of elements within a layer.

```{r}
p <- ggplot(data = diamonds) + aes(x=clarity, fill=cut)
plot1 <- p + geom_bar(position="fill") 
plot2 <- p + geom_bar(position="dodge") 
plot3 <- p + geom_bar(position="stack")
grid.arrange(plot1, plot2, plot3, ncol=3)
```

### Stats

It's possible to connect geoms to other statistics. Values computed by statistics are surrounded by double points, like `..count..`

```{r, warning=FALSE}
p <- ggplot(diamonds, aes(carat)) + xlim(0, 3)
plot1 <- p + stat_bin(aes(ymax = ..count..),   binwidth = 0.1, geom = "area")
plot2 <- p + stat_bin(aes(size = ..density..), binwidth = 0.1, geom = "point", position="identity")
grid.arrange(plot1, plot2, ncol=2)
```

```{r, warning=FALSE}
p <- ggplot(diamonds, aes(carat, price)) + xlim(1,3) 
plot1 <- p + stat_bin2d(bins = 25)
plot2 <- p + stat_binhex(bins = 10)
grid.arrange(plot1, plot2, ncol=2)
```

```{r, warning=FALSE}
p <- ggplot(diamonds, aes(carat, price)) + xlim(1,3) 
plot1 <- p + geom_point() + geom_density2d()
plot2 <- p + stat_density2d(geom = "point", aes(size = ..density..), contour = F) +
             scale_size_area(0.2, 1.5)
plot3 <- p + stat_density2d(geom = "tile", aes(fill = ..density..), contour = F)
grid.arrange(plot1, plot2, plot3, ncol=3)
```

### Annotating plots

```{r}
presdts <- presidential[-(1:3),] # remove the first 3 presidents
presdts$start <- as.Date(presdts$start)
head(presdts,4)
head(economics,4)
```

```{r}
p <- qplot(date, unemploy, data=economics, geom="line", 
           xlab = "", ylab = "No. unemployed (1000s)")

p + geom_vline(aes(xintercept = as.numeric(start)), data = presdts) +
    scale_x_date()

yrng <- range(economics$unemploy)
xrng <- range(economics$date)

library(scales) # use: alpha()
p + geom_rect(aes(NULL, NULL, xmin = start, xmax = end, fill = party), 
              ymin = yrng[1], ymax = yrng[2], data = presidential) +
    scale_fill_manual(values = alpha(c("blue", "red"), 0.2))

highest <- subset(economics, unemploy == max(unemploy))
p + geom_point(data = highest, size = 3, colour = alpha("red", 0.5))

p + geom_text(aes(x, y, label = "Unemployment rates of the last 40 years"), 
              data = data.frame(x = xrng[2], y = yrng[2]), 
              hjust = 1, vjust = 1, size = 4)
```

### Transformations

> Transformations carried out by the coordinate system change the appearance of the geoms: in polar coordinates a rectangle becomes a slice of a doughnut; in a map
projection, the shortest path between two points will no longer be a straight line -- Wickham

```{r, warning=FALSE}
p <- ggplot(data=data.frame(x=c(1,200),y=c(1,100))) + aes(x,y)
plot1 <- p + geom_hline(yintercept=seq(20, 60, by=15)) +
             geom_rect(xmin=20, xmax=40, ymin=20, ymax=60)
             
plot2 <- plot1 + coord_polar()              # x position mapped to angle
plot3 <- plot1 + coord_polar(theta="y")     # y position mapped to angle
plot4 <- plot1 + coord_flip()               # flip coordinates
plot5 <- plot1 + coord_equal()  
plot6 <- plot1 + coord_trans(x = "log10")
grid.arrange(plot1, plot2, plot3, plot4, plot5, plot6, nrow=2, ncol=3)
```

Eg with map coordinates:

```{r, message=FALSE}
library("maps")

m <- map_data("italy")
p <- ggplot(m, aes(x=long, y=lat, group=group)) +
            geom_polygon(fill="white", colour="black")
plot1 <- p                # Use cartesian coordinates
plot2 <- p + coord_map()  # With default mercator projection
grid.arrange(plot1, plot2, ncol=2)
```

### Some more complex egs

```{r, include=FALSE}
## Summarizes data.
## Gives count, mean, standard deviation, standard error of the mean, and confidence interval (default 95%).
##   data: a data frame.
##   measurevar: the name of a column that contains the variable to be summariezed
##   groupvars: a vector containing names of columns that contain grouping variables
##   na.rm: a boolean that indicates whether to ignore NA's
##   conf.interval: the percent range of the confidence interval (default is 95%)
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
    library(plyr)

    # New version of length which can handle NA's: if na.rm==T, don't count them
    length2 <- function (x, na.rm=FALSE) {
        if (na.rm) sum(!is.na(x))
        else       length(x)
    }

    # This does the summary. For each group's data frame, return a vector with
    # N, mean, and sd
    datac <- ddply(data, groupvars, .drop=.drop,
      .fun = function(xx, col) {
        c(N    = length2(xx[[col]], na.rm=na.rm),
          mean = mean   (xx[[col]], na.rm=na.rm),
          sd   = sd     (xx[[col]], na.rm=na.rm)
        )
      },
      measurevar
    )

    # Rename the "mean" column    
    datac <- rename(datac, c("mean" = measurevar))

    datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean

    # Confidence interval multiplier for standard error
    # Calculate t-statistic for confidence interval: 
    # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
    ciMult <- qt(conf.interval/2 + .5, datac$N-1)
    datac$ci <- datac$se * ciMult

    return(datac)
}
```

Let's see an example with error bars (`summarySE` is defined in the R markdown)):

```{r}
# from http://www.cookbook-r.com/Graphs/Plotting_means_and_error_bars_(ggplot2)/
tgc <- summarySE(ToothGrowth, measurevar="len", groupvars=c("supp","dose"))
head(tgc)

pd <- position_dodge(0.1) # move errorbars to the left and right

ggplot(tgc, aes(x=dose, y=len, colour=supp, group=supp)) + 
    geom_errorbar(aes(ymin=len-se, ymax=len+se), colour="black", width=.1, position=pd) +
    geom_line(position=pd) +
    geom_point(position=pd, size=3, shape=21, fill="white") + # 21 is filled circle
    xlab("Dose (mg)") +
    ylab("Tooth length") +
    scale_colour_hue(name="Supplement type",    # Legend label, use darker colors
                     breaks=c("OJ", "VC"),
                     labels=c("Orange juice", "Ascorbic acid"),
                     l=40) +                    # Use darker colors, lightness=40
    ggtitle("The Effect of Vitamin C on\nTooth Growth in Guinea Pigs") +
    expand_limits(y=0) +                        # Expand y range
    scale_y_continuous(breaks=0:20*4) +         # Set tick every 4
    theme_bw() +
    theme(legend.justification=c(1,0),
          legend.position=c(1,0))               # Position legend in bottom right
```

Eg of a population pyramid taken from Kyle Walker's  [example](https://rpubs.com/walkerke/pyramids_ggplot2):

```{r, include=FALSE}
library(XML)
library(reshape2)
library(plyr)


get_data <- function(country, year) {
  c1 <- "http://www.census.gov/population/international/data/idb/region.php?N=%20Results%20&T=10&A=separate&RT=0&Y="  
  c2 <- "&R=-1&C="
  url <- paste0(c1, year, c2, country)
  df <- data.frame(readHTMLTable(url))
  keep <- c(2, 4, 5)
  df <- df[,keep]  
  names(df) <- c("Age", "Male", "Female")
  cols <- 2:3
  df[,cols] <- apply(df[,cols], 2, function(x) as.numeric(as.character(gsub(",", "", x))))
  df <- df[df$Age != 'Total', ]  
  df$Male <- -1 * df$Male
  df$Age <- factor(df$Age, levels = df$Age, labels = df$Age)
  
  df.melt <- melt(df, 
                   value.name='Population', 
                   variable.name = 'Gender', 
                   id.vars='Age' )
  
  return(df.melt)
}

portugal <- get_data("PO", 2014)
```

```{r, warning=FALSE}
# males are counted with negative numbers, while females are with positive numbers
head(portugal) # check R markdown for code to produce this data frame

ggplot(portugal, aes(x = Age, y = Population, fill = Gender)) + 
  geom_bar(subset = .(Gender == "Female"), stat = "identity") + 
  geom_bar(subset = .(Gender == "Male"),   stat = "identity") + 
  scale_y_continuous(breaks = seq(-5e5, 5e5, 1.25e5), 
                     labels = paste0(as.character(c(seq(0.5, 0, -.125), seq(0.125, 0.5, .125))), "m")) + 
  coord_flip() +   # rotates 90º
  scale_fill_brewer(palette = "Set1") + 
  theme_bw()
```

## Using with maps

This next eg in taken from *ggmap, Spatial Visualization with ggplot2* vignette:

```{r, warning=FALSE, message=FALSE}
library(ggmap)

head(crime)  # compiled from Houston Police Department's website
murders <- crime[crime$offense == "murder",]

# get map of Houston (this is also a ggplot object)
houstonMap <- qmap("houston", zoom = 11, color = "bw", legend = "topleft")

plot1 <- houstonMap +
          geom_point(aes(x = lon, y = lat, colour = "red", size = 2), data = murders) +
          theme(legend.position="none")

plot2 <- houstonMap +
          stat_bin2d(aes(x = lon, y = lat, colour = offense, fill = offense),
                     size = .5, bins = 30, alpha = 1/2, data = murders)

plot3 <- houstonMap +
          stat_density2d(aes(x = lon, y = lat, fill = ..level.., alpha = ..level..),
                         size = 2, bins = 4, data = murders, geom = "polygon") +
          guides(alpha=FALSE) # remove alpha legend
grid.arrange(plot1, plot2, plot3, ncol=3)
```

This next eg is based on Robin Lovelace and James Cheshire's [tutorial](http://www.r-bloggers.com/introduction-to-spatial-data-and-ggplot2/) 
(the Rpubs page is [here](http://rpubs.com/m_dev/Intro-to-Spatial-Data-and-ggplot2)
and also [here](https://rpubs.com/RobinLovelace/ggmap)).

```{r, message=FALSE, message=FALSE}
library(rgdal)

sport <- readOGR(dsn = "files", "london_sport") # open files/london_sport
# sport.f <- fortify(sport, region = "ons_label") # format shapefile to plot
# head(sport.f)
# # add back attribute information associated with sport object (merge is a data join)
# sport.f <- merge(sport.f, sport@data, by.x = "id", by.y = "ons_label")
# head(sport.f)

sport <- SpatialPolygonsDataFrame(Sr = spTransform(sport, 
                                                   CRSobj = CRS("+init=epsg:4326")),
                                  data = sport@data)
sport.f  <- fortify(sport)    # format shapefile to be ploted
sport$id <- row.names(sport)  # provide same column names for join
sport.f  <- join(sport.f, sport@data)
head(sport.f)
```

With a well defined dataframe, gglopt can plot it using polygons (each London borough is called group):

```{r, fig.width=6, fig.height=6}
p <- ggplot(sport.f, aes(long, lat, group = group, fill = Partic_Per)) +
      geom_polygon() + 
      coord_equal() + 
      labs(x = "Easting (m)", y = "Northing (m)", fill = "% Sport Partic.") + 
      ggtitle("London Sports Participation")
p + scale_fill_gradient(low = "white", high = "black") # black&white version
# ggsave("plot.png", scale = 3, dpi = 400) # to save image
```

Several operations over the London map:

```{r}
p <- ggplot() + 
       geom_polygon(data = sport.f, aes(x = long, y = lat, group = group)) +
       coord_map() # this line of code ensures the plot is to scale
plot1 <- p
plot2 <- p + geom_point(aes(x = coordinates(sport)[, 1], y = coordinates(sport)[,2]))
plot3 <- plot2 + 
  geom_path(data = sport.f, aes(x = long, y = lat, group = group), color = "white") + 
  theme_classic()  # this line removes the distracting grey background

new_theme <- theme(axis.line = element_blank(), axis.ticks = element_blank(), 
                   axis.title.x = element_blank(), axis.title.y = element_blank(),
                   axis.text.y = element_blank(), axis.text.x = element_blank(),
                   axis.text.y = element_blank(), axis.text.y = element_blank(), 
                   panel.background = element_rect(fill = "lightgreen"))
plot4 <- p + new_theme
grid.arrange(plot1, plot2, plot3, plot4, nrow=2, ncol=2)
```

These objects can also be layered in ggmap objects:

```{r, warning=FALSE, fig.height=8, message=FALSE}
library(ggmap) 
b <- bbox(sport)
p <- ggmap(get_map(location = b))
plot1 <- p
plot2 <- p + 
  geom_polygon(data=sport.f, aes(x=long, y=lat, group=group, fill=Partic_Per), alpha=0.5) + 
  scale_fill_continuous(low = "green", high = "red") 

# using a new map source
p <- ggmap(get_map(location = b, source = "stamen", maptype = "toner", crop = T))
plot3 <- p + geom_polygon(data = sport.f, 
                     aes(x=long, y=lat, group=group, fill=Partic_Per), alpha = 0.5)
grid.arrange(plot1, plot2, plot3, nrow=2, ncol=2)
```



