---
title: "ggvis"
author: "João Neto"
date: October 2014
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 6
cache: yes
---

Refs:

+ [http://cran.r-project.org/web/packages/ggvis/vignettes/overview.html](http://cran.r-project.org/web/packages/ggvis/vignettes/overview.html)


> package ggvis is used to turn a dataset into a visualisation, setting up default mappings between variables in the dataset and visual properties. It uses a syntax similar in spirit to ggplot2 and dplyr.

> The goal is to combine the best of R (e.g. every modelling function you can imagine) and the best of the web (everyone has a web browser). Data manipulation and transformation are done in R, and the graphics are rendered in a web browser, using Vega [ref.(http://cran.r-project.org/web/packages/ggvis/vignettes/overview.html)

```{r, warning=FALSE, message=FALSE}
library(ggvis)
library(dplyr)
```

Several examples:

```{r}
head(mtcars)
mtcars %>% 
  ggvis(~wt, ~mpg) %>% 
  layer_points()
mtcars %>% 
  ggvis(~wt, ~mpg) %>% 
  layer_points(size := 25, shape := "diamond", stroke := "red", fill := NA)
mtcars %>% 
  ggvis(~mpg, ~disp, size = ~vs) %>% 
  layer_points()
mtcars %>% 
  ggvis(~mpg, ~disp, size = ~vs, opacity := 0.4) %>% 
  layer_points()
```

### Axis

```{r}
mtcars %>% 
  ggvis(~wt, ~mpg) %>% 
  layer_points() %>%
  add_axis("x", title = "Weight (lb/1000)", title_offset = 50, orient = "top") %>%
  add_axis("y", title = "Miles/(US) gallon", subdivide = 2, values = seq(10,34,by=2)) 

mtcars %>% 
  ggvis(~wt, ~mpg) %>% 
  layer_points() %>%
  add_axis("x", title = "") %>%
  add_axis("x", title = "Weight (lb/1000)", offset = 60, grid = FALSE) %>%
  hide_axis("y")

# For axes, you can set the properties of the ticks (or majorTicks and minorTicks separately), the labels and axis.
mtcars %>% ggvis(~wt, ~mpg) %>%
  layer_points() %>%
  add_axis("x", properties = axis_props(
    axis = list(stroke = "red", strokeWidth = 3),
    grid = list(stroke = "blue"),
    ticks = list(stroke = "blue", strokeWidth = 2),
    labels = list(angle = 45, align = "left", fontSize = 10)
  ))

mtcars %>% ggvis(y = ~mpg) %>%
  layer_points(prop("x", ~disp, scale = "xdisp")) %>%
  layer_points(prop("x", ~wt, scale = "xwt"), fill := "blue") %>%
  add_axis("x", "xdisp", orient = "bottom") %>%
  add_axis("x", "xwt", orient = "top",
    properties = axis_props(labels = list(fill = "blue")))
```

### Legends

Unlike ggplot2, by default, ggvis will not combine scales based on the same underlying variables into a single legend. Instead you must do this yourself by supplying the name of multiple scales to one legend:

```{r}
mtcars %>% 
  ggvis(~wt, ~mpg, size = ~cyl) %>% 
  layer_points() %>%
  add_legend(c("size"))
mtcars %>% 
  ggvis(~wt, ~mpg, size = ~cyl, fill=~cyl) %>% 
  layer_points() %>%
  add_legend(c("size","fill"))
```

### Regression Lines

```{r}
mtcars %>% 
  ggvis(~wt, ~mpg) %>%
  layer_points() %>%
  layer_smooths(span=0.5)  # add a smooth loess curve using span% of the neighbors
mtcars %>% 
  ggvis(~wt, ~mpg) %>%
  layer_points() %>%
  layer_model_predictions(model = "lm", se = TRUE, stroke := "red") # add a regression line with 95% CI
# layer_smooths() is just compute_smooth() + layer_paths()
mtcars %>%
  ggvis(~wt, ~mpg) %>%
  layer_points() %>%
  compute_smooth(mpg ~ wt) %>%
  layer_paths(~pred_, ~resp_, strokeWidth := 1, stroke := "red")
mtcars %>% ggvis(~wt, ~mpg) %>%
  layer_points(size := 5) %>%
  layer_smooths(span = 1, stroke := "orange") %>%
  layer_smooths(span = 0.3, stroke := "red")
```

### Interactive elements

Behind the scenes, interactive plots are built with shiny, and you can currently only have one running at a time in a given R session. To finish with a plot, press the stop button in Rstudio, or close the browser window and then press Escape or Ctrl + C in R.

```{r}
mtcars %>% 
  ggvis(~wt, ~mpg, 
    size := input_slider(10, 100, label = "Point size"),
    opacity := input_slider(0, 1, label = "Opacity"),
    stroke := input_select(c("GOLD" = "gold", "BLACK" = "black"), label="Select color"),
    strokeWidth := input_numeric(label = "Stroke size", value = 5)
  ) %>% 
  layer_points()
mtcars %>% 
  ggvis(~wt, ~mpg) %>% 
  layer_points() %>%
  layer_smooths(se = TRUE, span = input_slider(0.2, 1), stroke := "blue", strokeWidth := 5)
mtcars %>% ggvis(~wt, ~mpg) %>% layer_points() %>%
  layer_model_predictions(model = "lm",
    domain = input_slider(0, 10, value = c(1, 4)))  # domain determines the interval of the regression
mtcars %>% ggvis(~wt, ~mpg) %>% 
  layer_points() %>% 
  add_tooltip(function(df) df$wt*100)
```

### Scatter plots with grouping

```{r}
mtcars %>% 
  ggvis(~wt, ~mpg) %>% 
  layer_points(fill = ~factor(cyl))
mtcars %>% 
  ggvis(~wt, ~mpg, fill = ~factor(cyl)) %>% 
  layer_points() %>% 
  group_by(cyl) %>% 
  layer_smooths(span = input_slider(0.66, 1, value=1))
```

### Bar graphs

```{r}
head(pressure)
pressure %>% 
  ggvis(~temperature, ~pressure) %>%
  layer_bars(width = 12)
```

### Line graphs

```{r}
pressure %>% 
  ggvis(~temperature, ~pressure) %>% 
  layer_lines()
pressure %>% 
  ggvis(~temperature, ~pressure) %>%
  layer_points(shape := "square", stroke := "darkgreen", fill := "gold") %>% 
  layer_lines(stroke := "blue", strokeWidth := 1)
```

### Histograms

```{r}
head(faithful)
faithful %>% 
  ggvis(~eruptions) %>% 
  layer_histograms()
faithful %>% 
  ggvis(~eruptions) %>% 
  layer_histograms(width=0.5, center=0)
mtcars %>% 
  ggvis(~wt) %>% 
  layer_histograms(width =  input_slider(0, 2, step = 0.10, label = "width"),
                   center = input_slider(0, 2, step = 0.05, label = "center"))
cocaine %>% ggvis(~month, fill := "#fff8dc") %>%
  layer_histograms(width = 1, center = 0) %>%
  add_axis("x", title = "month") %>%
  add_axis("y", title = "count")
# with grouping
mtcars %>% 
  ggvis(~mpg, fill = ~factor(cyl)) %>% 
  group_by(cyl) %>%
  layer_histograms(width = 2, stack=TRUE)
```

### Box plots

```{r}
mtcars %>% 
  ggvis(~factor(cyl), ~mpg) %>% 
  layer_boxplots()
```

### More Layers

```{r}
mtcars %>% 
  arrange(wt) %>% 
  ggvis(~wt, ~mpg) %>%   
  layer_paths() # arrange+layer_paths is the same as layer_lines
t <- seq(0, 2 * pi, length = 100)
data.frame(x = sin(t), y = cos(t)) %>% 
  ggvis(~x, ~y) %>% 
  layer_paths(fill := "red") # with fill makes a polygon
mtcars %>% 
  arrange(wt) %>% 
  ggvis(~wt, ~mpg) %>%   
  layer_ribbons()
data.frame(x = 1:10, y = (1:10) ^ 2) %>%
  ggvis(~x, ~y, y2 := 0) %>% 
  layer_ribbons()
set.seed(1014)
data.frame(x1 = runif(5), x2 = runif(5), y1 = runif(5), y2 = runif(5)) %>% 
  ggvis(~x1, ~y1, x2 = ~x2, y2 = ~y2, fillOpacity := 0.1) %>% 
  layer_rects()
data.frame(x = 3:1, y = c(1, 3, 2), label = c("a", "b", "c"), type=c("x","x","y")) %>% 
  ggvis(~x, ~y, text := ~label) %>% 
  layer_text(fontSize := 15, fontWeight := "bold", angle := 45) %>%
  layer_text(text := ~type, dx := 10, dy := -10, stroke := "blue")
```

