---
title: "Fitting Distributions"
author: "João Neto"
date: October 2015
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
cache: TRUE
---

Refs:

+ [fitdistrplus: An R Package for Fitting Distributions](http://www.jstatsoft.org/article/view/v064i04/v64i04.pdf)

Introduction
===========

The typical way to fit a distribution is to use function `MASS::fitdistr`:

```{r}
library(MASS)
set.seed(101)
my_data <- rnorm(250, mean=1, sd=0.45)      # unkonwn distribution parameters

fit <- fitdistr(my_data, densfun="normal")  # we assume my_data ~ Normal(?,?)
fit
hist(my_data, pch=20, breaks=25, prob=TRUE, main="")
curve(dnorm(x, fit$estimate[1], fit$estimate[2]), col="red", lwd=2, add=T)
```

`fitdistr` uses `optim` to estimate the parameter values by maximizing the likelihood function. So it works like this:

```{r}
log_likelihood <- function(params) { -sum(dnorm(my_data, params[1], params[2], log=TRUE)) }
fit2 <- optim(c(0,1), log_likelihood)    # c(0,1) are just initial guesses
fit2
hist(my_data, pch=20, breaks=25, prob=TRUE)
curve(dnorm(x, fit2$par[1],     fit2$par[2]),     col="blue", lwd=6, add=T) # optim fit
curve(dnorm(x, fit$estimate[1], fit$estimate[2]), col="red",  lwd=2, add=T) # fitdistr fit
```

Using `fitdistrplus`
=========

This tutorial uses the `fitdistrplus` package for fitting distributions.

```{r, message=FALSE}
library(fitdistrplus)
```


Choosing which distribution to fit
-------------

Let's use one of the datasets provided by the packge:

```{r}
data("groundbeef", package = "fitdistrplus")
my_data <- groundbeef$serving
plot(my_data, pch=20)
```

We can first plot the empirical density and the histogram to gain insight of the data:

```{r, fig.width=10}
plotdist(my_data, histo = TRUE, demp = TRUE)
```

Another tool is to show some descriptive statistics, like the moments, to help us in making a decision:

```{r, fig.height=6, fig.width=6}
descdist(my_data, discrete=FALSE, boot=500)
```

The plot also includes a nonparametric bootstrap procedure for the values of kurtosis and skewness.

Fitting a distribution
-------------

Say, in the previous eg, we chose the weibull, gamma and log-normal to fit:

```{r}
fit_w  <- fitdist(my_data, "weibull")
fit_g  <- fitdist(my_data, "gamma")
fit_ln <- fitdist(my_data, "lnorm")
summary(fit_ln)
```

we can plot the results:

```{r, fig.height=10, fig.width=10}
par(mfrow=c(2,2))
plot.legend <- c("Weibull", "lognormal", "gamma")
denscomp(list(fit_w, fit_g, fit_ln), legendtext = plot.legend)
cdfcomp (list(fit_w, fit_g, fit_ln), legendtext = plot.legend)
qqcomp  (list(fit_w, fit_g, fit_ln), legendtext = plot.legend)
ppcomp  (list(fit_w, fit_g, fit_ln), legendtext = plot.legend)
```

```{r, echo=FALSE, eval=FALSE}
par(mfrow=c(1,1))
```

The fitting can work with other non-base distribution. It only needs that the correspodent, d, p, q functions are implemented.

In the next eg, the `endosulfan` dataset cannot be properly fit by the basic distributions like the log-normal:

```{r}
data("endosulfan", package = "fitdistrplus")
my_data <- endosulfan$ATV

fit_ln <- fitdist(my_data, "lnorm")
cdfcomp(fit_ln, xlogscale = TRUE, ylogscale = TRUE)
```

To solve this it is used the Burr and Pareto distributions available at package `actuar`

```{r, message=FALSE, fig.width=10}
library(actuar)

fit_ll <- fitdist(my_data, "llogis", start = list(shape = 1, scale = 500))
fit_P  <- fitdist(my_data, "pareto", start = list(shape = 1, scale = 500))
fit_B  <- fitdist(my_data, "burr",   start = list(shape1 = 0.3, shape2 = 1, rate = 1))
cdfcomp(list(fit_ln, fit_ll, fit_P, fit_B), xlogscale = TRUE, ylogscale = TRUE,
        legendtext = c("lognormal", "loglogistic", "Pareto", "Burr"), lwd=2)
```

The package also provides some goodness-of-it statistics:

```{r}
gofstat(list(fit_ln, fit_ll, fit_P, fit_B), fitnames = c("lnorm", "llogis", "Pareto", "Burr"))
```

Visually and using the previous statistics, it seems that the Burr distribution seems the preferred one among the candidates we chose to explore.

Parameter estimates
-----------

We can apply a bootstrap to estimate the uncertainty in the parameters:

```{r}
ests <- bootdist(fit_B, niter = 1e3)
summary(ests)
plot(ests)
quantile(ests, probs=.05) # 95% percentile bootstrap confidence interval
```




