---
title: "SQL Tutorial"
author: "João Neto"
date: "Thursday, July 31, 2014"
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
---

<!-- YAML options (above) -->

Preliminaries
-------------

Herein we are using the open-source database postgreSQL (more details [below](#postgreSQL)).

How to access a postgreSQL database using R:

```{r, warning=FALSE, message=FALSE}
library(RPostgreSQL)
 
drv <- dbDriver("PostgreSQL")
db <- dbConnect(drv, dbname="dvdrental", user="postgres", password="admin")
```

And to execute a query over the chosen database:
 
```{r}
query <- "SELECT
            *
          FROM
            customer;"
 
data <- dbGetQuery(db, query)
head(data)
```

SQL Language
------------

So SQL is a declarative language that specify relational algebra operations over tables (where rows represent records, and columns record attributes) in a relational database. Check [wikipedia](http://en.wikipedia.org/wiki/SQL) for more info.

This exercises are from [http://sqlschool.modeanalytics.com/](http://sqlschool.modeanalytics.com/)

### Selection

A query is the SQL way to retrieve information from one or more tables.

This is an eg of projection, ie, selection of just some columns:

```{r}
query <- "SELECT
            first_name, email
          FROM
            customer;"
 
data <- dbGetQuery(db, query)
head(data)
```

Rename the resulting columns:

```{r}
query <- 'SELECT
            first_name AS "FIRST NAME", 
            email AS "EMAIL"
          FROM
            customer;'
 
data <- dbGetQuery(db, query)
head(data)
```

Fix a limit on the queried data:

```{r}
query <- "SELECT
            first_name, email
          FROM
            customer
          LIMIT 3;"
 
data <- dbGetQuery(db, query)
head(data)
```

Another relational algebra operation is a filter:

```{r}
query <- "SELECT
            first_name, address_id, store_id
          FROM
            customer
          WHERE
            (address_id > 15 AND first_name <= 'm') OR store_id != 1;"
 
data <- dbGetQuery(db, query)
head(data,10)
```

The logical operators are `AND`, `OR` and `NOT`

To mix two columns:

```{r}
query <- "SELECT
            customer_id, 
            store_id, 
            1000 * customer_id + store_id AS mix_id
          FROM
            customer;"
 
data <- dbGetQuery(db, query)
head(data)
```

To make sorts:

```{r}
query <- "SELECT
            first_name, 
            last_name
          FROM
            customer
          ORDER BY
            first_name DESC, last_name;" # first name by decreasing order, then last name by increasing order
 
data <- dbGetQuery(db, query)
head(data,7)
```

Using (simple) regular expressions:

```{r}
query <- "SELECT
            first_name, 
            last_name
          FROM
            customer
          WHERE
            first_name LIKE 'W%a' OR last_name LIKE 'J_n%';"   # % means 0+ chars, _ means 1 char
 
data <- dbGetQuery(db, query)
head(data)
```

`ILIKE` is for case insensitive.

`IN` is used to compare with a set of values:

```{r}
query <- "SELECT
            first_name,    -- btw, this is a SQL comment
            last_name
          FROM
            customer
          /* 
                this is a multiline comment
           */
          WHERE
            first_name IN ('Mary', 'Barbara');" 
 
data <- dbGetQuery(db, query)
head(data)
```

The next egs use this google stock [csv](stocks.csv). To include it in postgreSQL do:

1. `CREATE DATABASE stocks;`
2. `\c stocks;`
3. `CREATE TABLE stocks (
    DATE_STOCK varchar(8), 
    YEAR integer,
    MONTH integer,
    OPEN double precision,
    HIGH double precision,
    LOW  double precision,
    CLOSE  double precision,
    VOLUME integer,
    id integer PRIMARY KEY
  );`
4. `\COPY stocks FROM 'C:\Users\jpn\Downloads\stocks.csv' DELIMITER ',' CSV HEADER;`  (adjust to your local dir)

Now we can access it here:

```{r}
db2 <- dbConnect(drv, dbname="stocks", user="postgres", password="admin")

query <- "SELECT
            *
          FROM
            stocks;" 
 
data <- dbGetQuery(db2, query)
head(data,10)
```

We can aggregate information into some available functions:

```{r}
query <- "SELECT
            COUNT(*),                 -- count the number of records
            COUNT(high) AS highs,     -- count the number of not null values of column 'high'
            SUM(high)   AS sum_highs, -- null values are treated as zero
            MIN(volume) AS min_volume,
            MAX(volume) AS max_volume,
            AVG(low)    AS avg_low    -- ignore nulls
          FROM
            stocks;" 
 
data <- dbGetQuery(db2, query)
head(data,10)
```

We can mix the aggregators with grouping the table into values:

```{r}
query <- "SELECT
            year,
            COUNT(*),                 -- count the number of records
            COUNT(high) AS highs,     -- count the number of not null values of column 'high'
            SUM(high)   AS sum_highs, -- null values are treated as zero
            MIN(volume) AS min_volume,
            MAX(volume) AS max_volume,
            AVG(low)    AS avg_low    -- ignore nulls
          FROM
            stocks
          GROUP BY
            year;" 
 
data <- dbGetQuery(db2, query)
head(data,10)
```

Groups can be nested:

```{r}
query <- "SELECT
            year, month,
            COUNT(*),                 -- count the number of records
            COUNT(high) AS highs,     -- count the number of not null values of column 'high'
            SUM(high)   AS sum_highs, -- null values are treated as zero
            MIN(volume) AS min_volume,
            MAX(volume) AS max_volume,
            AVG(low)    AS avg_low    -- ignore nulls
          FROM
            stocks
          GROUP BY
            year, month
          ORDER BY
            year, month;" 
 
data <- dbGetQuery(db2, query)
head(data,10)
```

Distinct values can be selected:

```{r}
query <- "SELECT DISTINCT
            month
          FROM
            stocks;" 
 
data <- dbGetQuery(db2, query)
head(data,10)
```

With more than one column, SQL returns all the distinct tuples it finds:

```{r}
query <- "SELECT DISTINCT
            month, year
          FROM
            stocks;" 
 
data <- dbGetQuery(db2, query)
head(data)
```

We can use it with count

```{r}
query <- "SELECT 
            COUNT(DISTINCT month) -- how many distinct months?
          FROM
            stocks;" 
 
data <- dbGetQuery(db2, query)
head(data)
```

For the next egs let's use this [csv](rosters.csv).

```{r}
db3 <- dbConnect(drv, dbname="rosters", user="postgres", password="admin")

query <- "SELECT
            *
          FROM
            players;" 
 
data <- dbGetQuery(db3, query)
head(data,10)
```

To deal with conditional statements, SQL uses `CASE WHEN ... THEN ... ELSE ... END`

```{r}
query <- "SELECT
            player,
            year,
            CASE WHEN year = 'SR'   -- create a new column where only senior players have a 'yes' values
              THEN 'yes'
              ELSE NULL 
            END AS is_a_senior
          FROM
            players;" 
 
data <- dbGetQuery(db3, query)
head(data,10)
```

It's possible to have a sequence of conditionals:

```{r}
query <- "SELECT
            player,
            weight,
            CASE WHEN weight > 250 THEN 'over 250'
                 WHEN weight > 200 THEN '200-250'
                 WHEN weight > 175 THEN '175-200'
                                   ELSE 'under 175' 
            END AS weight_group
          FROM
            players;" 
 
data <- dbGetQuery(db3, query)
head(data,10)
```

Conditional statements can be mixed with aggregatorsto achieve more expression power:

```{r}
query <- "SELECT
            CASE WHEN year = 'FR' 
              THEN 'FR'
              ELSE 'Not FR' 
            END      AS year_group,
            COUNT(1) AS count    -- 1 means the first column of the select, aka year_group
          FROM
            players
          GROUP BY 1;" 
 
data <- dbGetQuery(db3, query)
head(data)
```

Another eg:

```{r}
query <- "SELECT
            CASE WHEN year = 'FR' THEN 'FR'
                 WHEN year = 'SO' THEN 'SO'
                 WHEN year = 'JR' THEN 'JR'
                 WHEN year = 'SR' THEN 'SR'
                                  ELSE 'No Year Data' 
            END      AS year_group,
            COUNT(1) AS count
          FROM
            players
          GROUP BY 1;" 
 
data <- dbGetQuery(db3, query)
head(data)
```

The previous eg now rotated, aka, pivoting:

```{r}
query <- "SELECT
            COUNT(CASE WHEN year = 'JR' THEN 1 ELSE NULL END) AS jr_count,
            COUNT(CASE WHEN year = 'FR' THEN 1 ELSE NULL END) AS fr_count,
            COUNT(CASE WHEN year = 'SR' THEN 1 ELSE NULL END) AS sr_count,
            COUNT(CASE WHEN year = 'SO' THEN 1 ELSE NULL END) AS so_count
          FROM
            players;" 
 
data <- dbGetQuery(db3, query)
head(data)
```

### Joins

Joins permit that two or more tables can be associated to recover common associated information.

In the next egs we use a 2nd table names `teams` from the same database ([csv](teams.csv)).

```{r}
query <- "SELECT
            *
          FROM
            teams;" 
 
data <- dbGetQuery(db3, query)
head(data)
```

Say we want to know which conference (in teams table) has the highest average weight (in players table):

```{r}
query <- "SELECT
            *                   -- this is the full table returned by join, not the solution
          FROM
            players JOIN teams  -- this is the new table, the joined table, where select is going to operate
            ON players.school_name = teams.school_name;" 
 
data <- dbGetQuery(db3, query)
head(data)


query <- "SELECT
            teams.conference    AS conference,
            AVG(players.weight) AS average_weight
          FROM
            players JOIN teams
            ON teams.school_name = players.school_name
          GROUP BY teams.conference
          ORDER BY AVG(players.weight) DESC;" 
 
data <- dbGetQuery(db3, query)
head(data)
```

This `JOIN` is an _inner join_, ie, it will only show rows that satisfy the `ON` condition. It is an intersection of the two tables (check _outer joins_ below).

If there are columns in different tables with the same names but different information, we should use `AS` to give them distinct names.

### Outer Join

These are are tables for the next egs:

```{r}
db4 <- dbConnect(drv, dbname="crunchbase", user="postgres", password="admin")

query <- "SELECT
            *                
          FROM
            companies;" 
 
data <- dbGetQuery(db4, query)
head(data)
nrow(data)

query <- "SELECT
            *                
          FROM
            acquisitions;" 
 
data <- dbGetQuery(db4, query)
head(data)
nrow(data)
```

There are three types of Outer Join:
+ Left Outer Join, which includes unmatched rows from the left table
+ Right Outer Join, which includes unmatched rows from the right table (ie, `T2 LEFT JOIN T1` produces the same results as `T1 RIGHT JOIN T2`)
+ Full Outer Join, which includes unmatched rows from both tables

First the inner join:

```{r}
query <- "SELECT
            companies.permalink            AS companies_permalink,
            companies.name                 AS companies_name,
            acquisitions.company_permalink AS acquisitions_permalink
          FROM companies JOIN acquisitions
               ON companies.permalink = acquisitions.company_permalink;" 
 
data <- dbGetQuery(db4, query)
head(data)
```

Now a left join which includes several rows from companies table that does not have an entry at the acquisitions table:

```{r}
query <- "SELECT
            companies.permalink            AS companies_permalink,
            companies.name                 AS companies_name,
            acquisitions.company_permalink AS acquisitions_permalink
          FROM companies LEFT JOIN acquisitions
               ON companies.permalink = acquisitions.company_permalink;" 
 
data <- dbGetQuery(db4, query)
head(data)
```

The next eg count the number of unique companies (not double-counting companies) and unique *acquired* companies by state. It does not include results for which there is no state data, and order by the number of acquired companies from highest to lowest.

```{r}
query <- "SELECT companies.state_code,
                 COUNT(DISTINCT companies.permalink)            AS unique_companies,
                 COUNT(DISTINCT acquisitions.company_permalink) AS unique_companies_acquired
          FROM companies LEFT JOIN acquisitions
              ON companies.permalink = acquisitions.company_permalink
          WHERE companies.state_code IS NOT NULL
          GROUP BY 1
          ORDER BY 3 DESC;" 
 
data <- dbGetQuery(db4, query)
head(data)
```

If you want to provide a default value for the missing attributes, check function `COALESCE`.

### Converting Data

Functions `LEFT` and `RIGHT` allow to snip $n$ chars from a string. `LENGTH` returns the size of the string.

```{r}
query <- "SELECT
            company_permalink,
            LEFT(company_permalink, 6)  AS left_permalink,
            RIGHT(company_permalink, 6) AS right_permalink,
            LENGTH(company_permalink) AS length
          FROM
            acquisitions;" 
 
data <- dbGetQuery(db4, query)
head(data)
```

`TRIM` allows to remove certain chars from the extremeties of the string. It has 3 args: first from where we should start, then a string of char to remove, then a FROM followed by which column to process:

```{r}
query <- "SELECT
            company_permalink,
            TRIM(leading '/e' FROM company_permalink) AS trim_begin,
            TRIM(trailing '/e' FROM company_permalink) AS trim_end,
            TRIM(both '/e' FROM company_permalink) AS trim_begin_and_end
          FROM
            acquisitions;" 
 
data <- dbGetQuery(db4, query)
head(data)
```

Other functions:
+ `POSITION`, `STRPOS`, to find chars within a string
+ `SUBSTR`, `LEFT`, `RIGHT`, to produce substrings
+ `CONCAT`, `||`, to concatenate
+ `UPPER`, `LOWER`, to make uppercase/lowercase string

For date manipulation let's recheck the second db:

```{r}
query <- "SELECT
            date_stock, id
          FROM
            stocks;"
 
data <- dbGetQuery(db2, query)
head(data,20)
```

The `date_stock` column is a string with some quirks in it (different sizes for different dates). How to convert it to date?

```{r}
query <- "SELECT
            id, 
            date_stock,
            CASE WHEN LENGTH(date_stock) = 6                                  -- months 1 to 9, days 1 to 9
                    THEN ('20' || RIGHT(date_stock, 2) || '-' || 
                                  LEFT(date_stock, 1) || '-' || 
                                  SUBSTR(date_stock, 3, 1))::date 
                 WHEN LENGTH(date_stock) = 8                                  -- months > 9, days > 9
                    THEN ('20' || RIGHT(date_stock, 2) || '-' || 
                                  LEFT(date_stock, 2) || '-' || 
                                  SUBSTR(date_stock, 4, 2))::date 
                 WHEN LENGTH(date_stock) = 7 AND SUBSTR(date_stock,2,1) = '/' -- months < 10, days > 9
                    THEN ('20' || RIGHT(date_stock, 2) || '-' || 
                                  LEFT(date_stock, 1) || '-' || 
                                  SUBSTR(date_stock, 3, 2))::date 
                    ELSE ('20' || RIGHT(date_stock, 2) || '-' ||              -- months > 9, days < 10
                                  LEFT(date_stock, 2) || '-' || 
                                  SUBSTR(date_stock, 4, 2))::date   -- notice the data cast (à lá Haskell)
            END AS clean_date
          FROM
            stocks;"
 
data <- dbGetQuery(db2, query)
head(data, 20)
```

The function `NOW()` provides the current time (as a timestamp datatype):

```{r}
query <- "SELECT
            date_stock, id, NOW() AS current_time
          FROM
            stocks;"
 
data <- dbGetQuery(db2, query)
head(data)
```

Timestamps have specific function to retrive their composite information:

```{r}
query <- "SELECT
            EXTRACT('year' FROM temp_table.current_time) AS current_year
          FROM
            ( SELECT    -- make a sub_query, in order to get the timestamp for the eg
                date_stock, id, NOW() AS current_time 
              FROM stocks
            ) AS temp_table;"
 
data <- dbGetQuery(db2, query)
head(data)
```

### Subqueries

The previous eg has a subquery, the construction of a temporary (named) table needed to get the desired result.

Here's another eg that shows the rows with the smaller date in the table (the result is not ok, since, as we seen, the dates do not follow a standard description, but let's assume that was not a problem here). In this case the temporary table was made in the `WHERE` clause:

```{r}
query <- "SELECT
            date_stock, id
          FROM
            stocks
          WHERE
            date_stock = ( SELECT min(date_stock) FROM stocks );"
 
data <- dbGetQuery(db2, query)
head(data)
```

We can also have 2+ subqueries and then join them:

```{r}
query <- "SELECT
            first_name, last_name
          FROM
            customer;"
 
data <- dbGetQuery(db, query)
head(data)

# Show customers names that have last names started by 'An' or 'Ar'

query <- "SELECT
            temp.first_name, temp.last_name
          FROM
          ( SELECT first_name, last_name FROM customer WHERE last_name LIKE 'An%' 
            UNION ALL
            SELECT first_name, last_name FROM customer WHERE last_name LIKE 'Ar%' 
          ) AS temp;"
 
data <- dbGetQuery(db, query)
head(data,20)
```

```{r, echo=FALSE}
dbDisconnect(db)
dbDisconnect(db2)
dbDisconnect(db3)
dbDisconnect(db4)
```


About postgreSQL
-------------------

<a name="postgreSQL"/>

+ To run postgreSQL in admin mode: `psql -Upostgres`
+ To create a new database: `CREATE DATABASE dvdrental;`
+ To add a database into `dvdrental` database (at shell): `pg_restore -U postgres -d dvdrental dvdrental.tar`

More info at [http://www.postgresqltutorial.com/](http://www.postgresqltutorial.com/)

Helpful postgreSQL commands ([ref](http://blog.jasonmeridth.com/posts/postgresql-command-line-cheat-sheet/)):

+ change to postgres user and open psql prompt: `sudo -u postgres psql postgres`
+ list databases: `postgres=# \l`
+ list roles: `postgres=# \du`
+ create role: `postgres=#CREATE ROLE demorole1 WITH LOGIN ENCRYPTED PASSWORD 'password1' CREATEDB;`
+ alter role: `postgres=#ALTER ROLE demorole1 CREATEROLE CREATEDB REPLICATION SUPERUSER;`
+ drop role: `postgres=#DROP ROLE demorole1;`
+ create database: `postgres=#CREATE DATABASE demodb1 WITH OWNER demorole1 ENCODING 'UTF8';`
+ grant privileges to new user: `GRANT ALL PRIVILEGES ON DATABASE demodb1 TO demorole1;`
+ drop database: `postgres=#DROP DATABASE demodb1;`
+ connect to database: `\c <databasename>`
+ list tables in connected database: `\dt`
+ list columns on table: `\d <tablename>`
+ backup database: `pg_dump <databasename> > <outfile>`
+ quit: `\q`
+ help commands: `\?`
+ help command: `\h COMMAND`
+ execute commands from file: `\i FILE`
+ execute command in shell: `\! COMMAND`

Operations besides select using `RPostgreSQL`

```{r, warning=FALSE, message=FALSE}
library(RPostgreSQL)
 
drv <- dbDriver("PostgreSQL")
db <- dbConnect(drv, dbname="tests", user="postgres", password="admin")

# delete if exists
if (dbExistsTable(db, "books"))
  dbSendQuery(db, "DROP TABLE books;") # or  dbRemoveTable("books");

# create table
dbSendQuery(db, "CREATE TABLE books (intcolumn integer, floatcolumn float);")

# insert data, row by row
dbSendQuery(db, "INSERT INTO books VALUES(12, 19.95);")
dbSendQuery(db, "INSERT INTO books VALUES( 1,  9.95);")
dbSendQuery(db, "INSERT INTO books VALUES(10, 14.95);")
dbSendQuery(db, "INSERT INTO books VALUES( 2,  4.95);")
dbSendQuery(db, "INSERT INTO books VALUES( 3,  1.45);")

# get contents
dbGetQuery(db, "SELECT * FROM books;")

# get contents (lazy)
data <- dbSendQuery(db, "SELECT * FROM books;")
fetch(data, 3)       # get the next 2 records (use -1 to fetch all)
dbGetRowCount(data)  # how many rows were processed so far
fetch(data, 2) 
dbGetRowCount(data)

# get the DBMS statement of a given resulst
dbGetStatement(data)
dbGetRowsAffected(data) # how many rows were affected (-1 if none was)

# get avalailable tables in database
dbListTables(db)

# get list of table's column names
dbListFields(db, "books")

# import table into data frame
df <- dbReadTable(db, "books")
df

# export data frame into db table
df2 <- data.frame(id=1:5, x=runif(10), y=sample(letters,10))
df2
dbWriteTable(db, "new_table", df2)
dbReadTable(db, "new_table")

dbClearResult(data) # flushes any pending data and frees the resources used by result set

# postgreSQL transactions
dbSendQuery(db, "BEGIN;")            # begin transaction
dbRemoveTable(db,"new_table")
dbExistsTable(db,"new_table")
dbRollback(db)                       # rollback, ie, undo all changes since last begin
dbExistsTable(db,"new_table")
dbSendQuery(db, "BEGIN;")
dbRemoveTable(db,"new_table")
dbExistsTable(db,"new_table")
dbCommit(db)                         # or dbSendQuery(db, "COMMIT;") 
dbExistsTable(db,"new_table")

dbDisconnect(db)    # disconnects with the database
dbUnloadDriver(drv) # unload driver (frees all resources)
```
