---
title: "Caret"
author: "João Neto"
date: "February 2015"
output: 
  html_document:
    toc: true
    toc_depth: 3
    fig_width: 12
    fig_height: 6
cache: TRUE
---
Ref:

+ [Short Intro to cater](http://cran.r-project.org/web/packages/caret/vignettes/caret.pdf)

+ [Caret Model Training and Tuning](http://topepo.github.io/caret/training.html)

> The caret package (short for Classification And REgression Training) is a set of functions that attempt to streamline the process for creating predictive models. The package contains tools for: data splitting
pre-processing
feature selection
model tuning using resampling
variable importance estimation
as well as other functionality. [ref](http://topepo.github.io/caret/training.html)

```{r}
# to install package with all dependencies:
# install.packages("caret", dependencies = c("Depends", "Suggests"))
library(caret)
```

To partition data into training and test sets:

```{r}
inTrain <- createDataPartition(y = iris$Species, # which columns is the outcome data
                               p = .75,          # training set relative size
                               list = FALSE)
training <- iris[ inTrain,]
testing  <- iris[-inTrain,]
```

Given some model, `train` tune it according to the data. Function `train` recognizes a [huge number of models](http://topepo.github.io/caret/bytag.html). To see the current list:

```{r}
names(getModelInfo())
```

For this first example, let's use support vector machines with linear kernels:

```{r, warning=FALSE, message=FALSE}
svm_fit <- train(Species ~ .,
                 data = training,
                 method = "svmLinear",
                 preProcess = c("center", "scale"))
svm_fit
```

And we use the trained model to make predictions:

```{r}
test_pred <- predict(svm_fit, newdata=testing)
confusionMatrix(data=test_pred, testing$Species)
```

It's possible to control how the model tuning is done:

```{r}
ctrl <- trainControl(method = "repeatedcv", # K-fold cross validation w/repeated train/test splits
                     number = 8,            # K value (default 10)
                     repeats = 3)           # number of repetitions (for repeatedcv only)

# if the tuning is done, method="none" will not use resampling nor parameter tuning

svm_fit <- train(Species ~ .,
                 data = training,
                 method = "svmLinear",
                 tuneLength = 15,  # how many candidate parameter values to evaluate
                 trControl = ctrl,
                 preProc = c("center", "scale"))
svm_fit

test_pred <- predict(svm_fit, newdata=testing)
confusionMatrix(data=test_pred, testing$Species)
```

In this case, we don't notice changes, since the initial training was already almost perfect.

There's also an option `tuneGrid` that allows the user to pass parameter values he chooses:

```{r}
tune_grid <-  expand.grid(interaction.depth = c(1, 5, 9),
                          n.trees = (1:30)*10,
                          shrinkage = 0.1)

gbm_fit <- train(Species ~ .,
                 data = training,
                 method = "gbm",
                 tuneGrid = tune_grid,
                 trControl = ctrl,
                 preProc = c("center", "scale"),
                 metric = "Kappa", # which metric to optimize ()
                 verbose=FALSE)

test_pred <- predict(gbm_fit, newdata=testing)
confusionMatrix(data=test_pred, testing$Species)

plot(gbm_fit)
plot(gbm_fit, metric="Kappa")
```

There are options to pre-process data (besides centering and scaling) that can be checked [here](http://topepo.github.io/caret/preprocess.html).

To compare different models:

```{r}
resamps <- resamples(list(GBM = gbm_fit,
                          SVM = svm_fit))
resamps
summary(resamps)
bwplot(resamps, layout = c(2, 1))
dotplot(resamps, metric = "Kappa")
xyplot(resamps, what = "BlandAltman")
splom(resamps)
```

A company package `caretEmsemble` deals with ensembles of models, and a vignette can be read [here](http://cran.r-project.org/web/packages/caretEnsemble/vignettes/caretEnsemble-intro.html).